---
title: "Exercises for Data Processing with R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, exercise.lines=5, exercise.eval=FALSE)
```


## Data Structures

### Exercise 1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span>.

```{r dsex1, exercise=TRUE}
mydf = 
# mymatrix = 
```


```{r dsex1-hint}
mydf = data.frame(A=1:3, B=letters[1:3])
str(mydf)
```


### Exercise 2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created.  You'll have to recreate the data.frame/matrix here.

```{r dsex2, exercise=TRUE, exercise.eval=TRUE}


```

```{r dsex2-hint}
mylist = list(c('a', 'b'), 1:3, mydf)
```



#### Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?




## I/O


### Exercise 1

Use <span class="">readr</span> and <span class="">haven</span> to read the following files. Use the url just like you would any filename.  The latter is a Stata file.  You can use the RStudio's menu approach to import the file if you want.


```{r ioex1, exercise=TRUE}
'https://raw.githubusercontent.com/m-clark/data-processing-and-visualization/master/data/cars.csv'

'https://raw.githubusercontent.com/m-clark/data-processing-and-visualization/master/data/presvote.dta'
```

```{r ioex1-hint}
readr::read_csv()
```


### Thinking Exercises

Why might you use <span class="func">read_csv</span> from the <span class="pack">readr</span> package rather than <span class="func">read.csv</span> in base R?

What is your definition of 'big' data?




## Indexing Exercises

This following is a refresher of base R indexing only.

Here is a <span class="objclass">matrix</span>, a <span class="objclass">data.frame</span> and a <span class="objclass">list</span>.

```{r ixex0}
mymatrix = matrix(rnorm(100), 10, 10)
mydf = cars
mylist = list(mymatrix, thisdf = mydf)
```

### Exercise 1

For the <span class="objclass">matrix</span>, in separate operations, take a slice of rows, a selection of columns, and a single element.

```{r ixex1, exercise=TRUE}
mymatrix = matrix(rnorm(100), 10, 10)

```

```{r ixex1-hint}
mymatrix[1:5, ]
mymatrix[, 1:5]
mymatrix[1,2]
```

### Exercise 2

For the <span class="objclass">data.frame</span>, grab a column in 3 different ways.

```{r ixex2, exercise=TRUE}
mydf = cars
```

```{r ixex2-hint}
mydf$disp
mydf[,2]
mydf['disp']
```

### Exercise 3

For the <span class="objclass">list</span> grab an element by number and by name.

```{r ixex3, exercise=TRUE}
mylist = list(mymatrix, thisdf = mydf)
```

```{r ixex3-hint}
mylist[2]
mylist$thisdf
```





## Vector/Apply Exercises


### Exercise 1

With the following <span class="objclass">matrix</span>, use <span class="func">apply</span> and the <span class="func">sum</span> function to get row or column sums.  

```{r vectorapp_ex1, exercise=TRUE}
x = matrix(1:9, 3, 3)
```

```{r vectorapp_ex1-hint}
apply(x, ???, sum)
```

### Exercise 2

With the following <span class="objclass">list</span>, use <span class="func">lapply</span> and <span class="func">sapply</span> and the <span class="func">sum</span> function to get sums for the elements.  There is no margin to specify with on a list, so just supply the list and the sum function.  

```{r vectorapp_ex2, exercise=TRUE}
x = list(1:3, 4:6, 7:9)
```

```{r vectorapp_ex2-hint}
lapply(x, sum)
sapply(x, sum)
```

<span class="func">sapply</span> is actually just a wrapper for <span class="func">lapply</span>.  If you supply the argument `simplified=F`, it is identical. Otherwise, it attempts to return a vector or matrix.


## dplyr Exercises

### Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r init_dplyr, exercise=TRUE, message=FALSE}
install.packages('ggplot2movies')
library(ggplot2movies)
data('movies')
```


### Exercise 1

Using the movies data set, perform each of the following actions separately.

#### Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```{r dpex1a, exercise=TRUE}
movies %>% 
  mutate(new_var_name = '?')
```

```{r dpex1a-hint}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

#### Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r dpex1b, exercise=TRUE, exercise.eval=FALSE}
movies %>% 
  filter(?) 
```

```{r dpex1b-hint}
movies %>% 
  filter(year >= 2000) 
```

#### Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the `title`, `year`, `budget`, `length`, `rating` and `votes` variables.  There are at least 3 ways to do this.

```{r dpex1c, exercise=TRUE, exercise.eval=FALSE}
movies %>% 
  select(?) 
```

```{r dpex1c-hint}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```



### Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.

Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r dpex2, exercise=TRUE, exercise.eval=FALSE}
movies %>%
  group_by(?) %>%
  summarise(?) %>%
  tail()
```

```{r dpex2-hint}
movies %>%
  group_by(year) %>%
  summarise(AvgBudget=mean(budget, na.rm=T)) %>%
  tail()
```


### Exercise 3

Now put it all together in one set of piped operations.

- Filter movies released *after* 1990
- select the same variables as before but also the `mpaa`, `Action`, and `Drama` variables
- group by `mpaa` *and* (your choice) `Action` *or* `Drama`
- get the average rating



```{r dpex3, exercise=TRUE, exercise.eval=FALSE}
movies %>%
  filter(?) %>%
  select(?) %>%
  group_by(?) %>%
  summarise(?)
```

```{r dpex3-hint}
movies %>%
  filter(year>=1990) %>%
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>%
  group_by(mpaa, Drama) %>%
  summarise(AvgRating = mean(rating))
```


It should spit out something like the following:

```{r dpex3_show, echo=FALSE, message=FALSE}
data('movies', package='ggplot2movies')
library(dplyr)
movies %>%
  filter(year>=1990) %>%
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>%
  group_by(mpaa, Drama) %>%
  summarise(AvgRating = mean(rating))
```





## data.table Exercises


### Exercise 0

Install and load the <span class="pack">data.table</span> package.



```{r dtex0, message=FALSE, exercise=TRUE, exercise.eval=FALSE}
install.packages('data.table')

```



### Exercise 1

Create the following data table. Then, create a new object that contains only the 'a' group. Think back to how you use a logical to select rows.

```{r dtex1, exercise=TRUE, exercise.eval=FALSE, message=FALSE}
library(data.table)

mydt = data.table(expand.grid(x=1:3, y=c('a','b','c')), z=sample(1:20, 9))

mydt_a = mydt[?,]
```

```{r dtex1-hint}
mydt_a = mydt[y=='a',]
mydt_a
```


### Exercise 2

Create a new object that is the sum of z grouped by x. You don't need to name the sum variable.


```{r dtex2, exercise=TRUE, exercise.eval=FALSE}
mydt = data.table(expand.grid(x=1:3, y=c('a','b','c')), z=sample(1:20, 9))

mydt_grpsum = mydt[,?]
```
```{r dtex2-hint}
mydt_grpsum = mydt[,sum(z), by=x]
mydt_grpsum = mydt[,list(sumz=sum(z)), by=x]
```
