# Data Structures

```{r dssetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=FALSE, cache.rebuild=F, cache=F,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
```


<!-- <div style="text-align:center; font-size:200%; font-variant:small-caps;" ><span style='color:#1e90ff'>Data Structures</span></div> -->
  
R has several core data structures:
  
- Vectors
- Factors
- Lists
- Matrices/arrays
- Data frames


### Vectors

<span class="objclass">Vectors</span> form the basis of R data structures.

Two main types- <span class="objclass">atomic</span> and <span class="objclass">lists</span>, but I will treat lists separately.

Here is an R <span class="objclass">vector</span>. The <span class="emph">elements</span> of the vector are numeric values.

```{r vec, eval=TRUE}
x = c(1, 3, 2, 5, 4)
x
```


All elements of an atomic vector are the same type. Examples include:
  
- <span class="objclass">character</span>
- <span class="objclass">numeric</span> (double)
- <span class="objclass">integer</span>
- <span class="objclass">logical</span>



### Factors

A important type of vector is a <span class="objclass">factor</span>. Factors are used to represent categorical data structures.

```{r factors, eval=TRUE}
x = factor(1:3, labels=c('q', 'V', 'what the heck?'))
x
```


The underlying representation is numeric, but it is important to remember that factors are *categorical*.

They can't be used as numbers would be, as the following demonstrates.

```{r factors2, eval=TRUE, error=TRUE}
as.numeric(x)
sum(x)
```



### Matrices

With multiple dimensions, we are dealing with <span class="objclass">arrays.</span>

<span class="objclass">Matrices</span> are 2-d arrays, and extremely commonly used for scientific computing.

The vectors making up a matrix *must all be of the same type*.

- e.g. all values in a matrix must be numeric.



#### Creating a matrix

Creating a matrix can be done in a variety of ways.

```{r createMatrix}
# create vectors
x = 1:4
y = 5:8
z = 9:12

rbind(x, y, z)   # row bind
cbind(x, y, z)   # column bind
matrix(c(x, y, z), nrow=3, ncol=4, byrow=TRUE)
```


### Lists

<span class="objclass">Lists</span> in R are highly flexible objects, and probably the most commonly used for data science.

They can contain anything as their elements, even other lists.

- unlike vectors, whose elements must be of the same type. 

Here is a list. We use the <span class="func">list</span> function to create it. 

```{r list, eval=TRUE}
x = list(1, "apple", list(3, "cat"))
x
```

We often want to loop some function over a list.

```{r listloop, eval=TRUE}
for(elem in x) print(class(elem))
```

Lists can, and often do, have named elements.

```{r namedlist, eval=TRUE}
x = list("a" = 25, "b" = -1, "c" = 0)
x["b"]
```



### Data Frames

<span class="objclass">data.frames</span> are a very commonly used data structure. They do not have to have the same type of element, and this is because the data.frame class is actually just a list.

- As such, everything about lists applies to them

But they can also be indexed by row or column as well.

- like matrices

There are other very common types of object classes associated with packages that are both a <span class="objclass">data.frame</span> and other type of structure.


#### Creating a data frame

```{r createdf, eval=TRUE}
mydf = data.frame(a = c(1,5,2), 
                  b = c(3,8,1))
```

We can add row names also.

```{r dfrownames, eval=TRUE}
rownames(mydf) = paste0('row', 1:3)
mydf
```


### Exercise

#### Excercise #1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span>.

```{r dsex1, echo=F}
mydf = data.frame(A=1:3, B=letters[1:3])
```


#### Exercise #2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created.

```{r dsex2, echo=F}
mylist = list(c('a', 'b'), 1:3, mydf)
```



#### Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?