# Indexing

```{r indexsetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=FALSE, cache.rebuild=F, cache=F,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
```


### Base R Indexing Refresher

##### Slicing vectors

```{r vectorSlice, eval=TRUE}
letters[4:6]
letters[c(13,10,3)]
```

##### Slicing matrices/data.frames

```{r matrixSlice}
myMatrix[1, 2:3]
```


##### Label-based indexing

```{r dfindexlabel}
mydf['row1', 'b']
```

##### Position-based indexing

```{r dfindexpos}
mydf[1, 2]
```


##### Mixed indexing

```{r dfindexmix}
mydf['row1', 2]
```

If the row/column value is empty, all rows/columns are retained.

```{r dfindexslice}
mydf['row1',]
mydf[,'b']
```


##### Non-contiguous

```{r dfindexnoncont}
mydf[c(1,3),]
```

##### Boolean

```{r dfindexbool}
mydf[mydf$a >= 2,]
```


### List/Data.frame extraction

<span class="func">[</span> : grab a slice of elements/columns

<span class="func">[[</span> : grab specific elements/columns

<span class="func">$</span> : grab specific elements/columns

<span class="func">@</span>: extract slot for S4 objects

```{r dflistslice}
my_list_or_df[2:4]
```


```{r dflistextract}
my_list_or_df[['name']]
```

```{r dflistextract2}
my_list_or_df$name
```


### Exercises

This following is a refresher of base R indexing only.

Here is a matrix, a data.frame and a list.

```{r}
mymatrix = matrix(rnorm(100), 10, 10)
mydf = cars
mylist = list(mymatrix, thisdf = mydf)
```

#### Exercise 1

For the matrix, in separate operations, take a slice of rows, a selection of columns, and a single element.

```{r, echo=F}
mymatrix[1:5, ]
mymatrix[, 1:5]
mymatrix[1,2]
```

#### Exercise 2

For the data.frame, grab a column in 3 different ways.

```{r, echo=F}
mydf$dist
mydf[,2]
mydf['dist']
```

#### Exercise 3

For the list grab an element by number and by name.

```{r, echo=FALSE}
mylist[2]
mylist$thisdf
```

