# Interactive Visualization

```{r title, echo=F, cache=F, eval=TRUE}
nodes = data.frame(id=1:2, label=c('Interactive', 'Visualization'), color='#1e90ff')
edges = data.frame(from=1, to=2, color='transparent')
library(visNetwork)
visNetwork(nodes, edges, height=400, width='100%') %>% 
  visNodes(shape='text', font=list(color='#1e90ff', size=164)) %>% 
  # visPhysics(solver='barnesHut', barnesHut=list(gravitationalConstant = -50000, springConstant=.11, springLength=50)) %>% 
  visOptions(highlightNearest=F) 
# %>%
  # visHierarchicalLayout(direction='LR')
```


### Packages

<span class="pack">ggplot2</span> is the most widely used package for visualization in R.

However, it is not interactive by default.

Many packages use <span class="pack">htmlwidgets</span>, **d3** (JavaScript library) etc. to provide interactive graphics.

General:

- <span class="pack">plotly</span>
  - used also in Python, Matlab, Julia, can convert ggplot2 images to interactive ones.
- <span class="pack">highchartr</span>
  - also very general wrapper for highcharts.js and works with some R packages out of the box
- <span class="pack">rbokeh</span>
  - like plotly, it also has cross program support

Specific functionality:

  - <span class="pack">DT</span>
    - interactive data tables
  - <span class='pack'>leaflet</span>
      - maps with OpenStreetMap
  - <span class='pack'>visNetwork</span>
      - Network visualization
  

### Piping for Visualization

One of the advantages to piping is that it's not limited to <span class="pack">dplyr</span> style data management functions.

<span class='emph'>*Any*</span> R function can be potentially piped to.

- several examples have already been shown.

This facilitates data exploration, especially visually.

- don't have to create objects
- new variables are easily created and subsequently manipulated just for vis
- data manipulation not separated from visualization


### htmlwidgets

Many newer visualization packages take advantage of piping.

<span class='pack'>htmlwidgets</span> is a package that makes it easy to create javascript visualizations.

- i.e. what you see everywhere on the web.

The packages using it typically are pipe-oriented and produce interactive plots.



### plotly example

A couple demonstrations with <span class="pack">plotly</span>.

Note the layering as with <span class="pack">ggplot2</span>.

Piping used before plotting.

```{r plotly1, cache=FALSE, eval=TRUE}
library(plotly)
midwest %>% 
  filter(inmetro==T) %>% 
  plot_ly(x=~percbelowpoverty, y=~percollege, mode='markers') 
```


<span class="pack">plotly</span> has <span class="emph">modes</span>, which allow for points, lines, text and combinations.

<span class="emph">Traces</span> work similar to geoms.

```{r plotly2, echo=TRUE, cache=FALSE, eval=TRUE}
library(mgcv); library(modelr)

mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor)) %>% 
  add_predictions(gam(mpg~s(wt, am, bs='fs'), data=mtcars)) %>% 
  arrange(wt) %>% 
  plot_ly(x=~wt, y=~mpg, color=~amFactor, mode='markers') %>% 
  add_trace(x=~wt, y=~pred, color=~amFactor, alpha=.5, hover=~hovertext, name='gam prediction')
```

### ggplotly

The nice thing about <span class="pack">plotly</span> is that we can feed a <span class="objclass">ggplot</span> to it.

It would have been easier to use <span class="func">geom_smooth</span>, so let's do so.


```{r ggplotly, cache=FALSE, eval=TRUE}
gp = mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor),
         prediction = predict(gam(mpg~s(wt), data=mtcars))) %>% 
  arrange(wt) %>% 
  ggplot(aes(x=wt, y=mpg)) +
  geom_smooth() +
  geom_point(aes(color=amFactor))
ggplotly()
```




### Highcharter

<span class="pack">Highcharter</span> is also fairly useful for a wide variety of plots.

- Uses the <span class="pack">highcharts.js</span> library

```{r dygraphdata, cache=FALSE, eval=F, echo=FALSE}
# library(dygraphs)
# data(UKLungDeaths)
# cbind(ldeaths, mdeaths, fdeaths) %>% 
#   dygraph() %>% 
#   dyOptions(stackedGraph = TRUE, colors=RColorBrewer::brewer.pal(3, name='Dark2')) %>%
#   dyRangeSelector(height = 20)
```

```{r highcharts, cache=FALSE, eval=T}
library(highcharter); library("quantmod")

x <- getSymbols("GOOG", auto.assign = FALSE)

highchart(type='stock', width=1000) %>% 
   hc_add_series(data = x) 
```

### visNetwork

<span class="pack">visNetwork</span> allows for network visualizations

- Uses the <span class="pack">vis.js</span> library

```{r visNetworkinitial, echo=FALSE, cache=TRUE, eval=T}
set.seed(1352)
nodes = data.frame(id = 0:5,
                   label = c('Bobby', 'Janie','Timmie', 'Mary', 'Johnny', 'Billy'),
                   group = c('friend', 'frenemy','frenemy', rep('friend', 3)),
                   value = sample(10:50, 6))
edges = data.frame(from = c(0,0,0,1,1,2,2,3,3,3,4,5,5),
                   to = sample(0:5, 13, replace = T),
                   value = sample(1:10, 13, replace = T)) %>% 
  filter(from!=to)
```

```{r visNetwork, cache=FALSE, eval=F, echo=-1}
saveWidget(vn, file='visnetwork.html')
library(visNetwork)
visNetwork(nodes, edges) %>% #, height=600, width=800
  visNodes(shape='circle', 
           font=list(), 
           scaling=list(min=10, max=50, label=list(enable=T))) %>% 
  visLegend()
```

<iframe src='../img/visnetwork.html', width=1000, height=600, scrolling="no", frameBorder="0"></iframe>


### data table

Use the <span class="pack">DT</span> package for interactive dataframes.

```{r datatable, cache=FALSE, eval=TRUE}
library(DT)
ggplot2movies::movies %>% 
  select(1:6) %>% 
  filter(rating>9) %>% 
  slice(sample(1:nrow(.), 50)) %>% 
  datatable(rownames=F)
```


### Shiny

<span class="emph">Shiny</span> is a framework that can essentially allow you to build an interactive website.

- Provided by RStudio developers

Most of the more recently developed visualization packages will work specifically within the shiny and rmarkdown settings.


### Interactive and Visual Data Exploration

Interactivity allows for even more dimensions to be brought to a graphic.

Interactive graphics are more fun too!

- But they must serve a purpose
- Too often they are simply distraction, and detract from the data story

Just a couple visualization packages can go a very long way.


