# Part 1: Data Processing

## Data Structures

R has several core data structures, and we'll take a look at each.
  
- Vectors
- Factors
- Lists
- Matrices/arrays
- Data frames

The more you know about R data structures, the more you'll know how to use them, the more you'll know *why* things go wrong if they do, and the further you'll be able to go with your data.

### Vectors

<span class="objclass">Vectors</span> form the basis of R data structures. Two main types are <span class="objclass">atomic</span> and <span class="objclass">lists</span>, but we'll talk about lists separately.

Here is an R <span class="objclass">vector</span>. The <span class="emph">elements</span> of the vector are numeric values.

```{r vec}
x = c(1, 3, 2, 5, 4)
x
```


All elements of an atomic vector are the same <span class="emph">type</span>. Example types include:
  
- <span class="objclass">character</span>
- <span class="objclass">numeric</span> (double)
- <span class="objclass">integer</span>
- <span class="objclass">logical</span>

In addition, there are special kinds of values like <span class="objclass">NA</span> ('not available' i.e. missing), <span class="objclass">NULL</span>, <span class="objclass">NaN</span> (not a number), <span class="objclass">Inf</span> (infinite) and so forth.

<div class='note'>
You can use <span class="func">typeof</span> to examine an object's type, or use an `is` function, e.g. <span class="func">is.logical</span>, to check if an object is a specific type.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

#### Character strings

When dealing with text, objects of class character are what you'd typically be dealing with.  

```{r create_a_char, eval=F}
x = c('... Of Your Fake Dimension', 'Ephemeron', 'Dryswch', 'Isotasy', 'Memory')
x
```

Not much to it, but be aware there is no real limit to what is represented as a character vector. For example, in a data frame, you could have a column where each entry is one of the works of Shakespeare.


#### Factors

An important type of vector is a <span class="objclass">factor</span>. Factors are used to represent categorical data structures.  Although not exactly precise, one can think of factors as integers with labels.  So the underlying representation of a variable for <span class="objclass">sex</span> is 1:2 with labels 'Male' and 'Female'.  They are a special class with attributes, or metadata, that contains the information about the <span class="objclass">levels</span>.

```{r factor_atts}
x = factor(rep(letters[1:3], e=10))
x
attributes(x)
```


```{r factors, eval=F, echo=FALSE}
x = factor(1:3, labels=c('q', 'V', 'what the heck?'))
x
```


The underlying representation is numeric, but it is important to remember that factors are *categorical*.  Thus, they can't be used as numbers would be, as the following demonstrates.

```{r factors2, error=TRUE}
x_num = as.numeric(x)  # convert to a numeric object
sum(x_num)
sum(x)
```

##### Strings vs. Factors

The main thing to note is that factors are generally a statistical phenomenon, and are required to do statistical things with data that would otherwise be a simple character string.  If you know the relatively few levels the data can take, you'll generally want to use factors, or at least know that statistical packages and methods may require them.  In addition, factors allow you to easily overcome the silly default alphabetical ordering of category levels in some very popular visualization packages.

For other things, such as text analysis, you'll almost certainly want character strings instead, and in many cases it will be required.  It's also worth noting that a lot of base R and other behavior will coerce strings to factors.  This made a lot more sense in the early days of R, but is not really necessary these days.


<div class='note'>
Some packages to note to help you with processing strings and factors:

- <span class="pack">forcats</span>
- <span class="pack">stringr</span>
<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

### Matrices

With multiple dimensions, we are dealing with <span class="objclass">arrays</span>. <span class="objclass">Matrices</span> are 2-d arrays, and extremely commonly used for scientific computing[^tidynomatrix]. The vectors making up a matrix *must all be of the same type*.  For example, all values in a matrix might be <span class="objclass">numeric</span>, or all <span class="objclass">character strings</span>.



#### Creating a matrix

Creating a matrix can be done in a variety of ways.

```{r createMatrix}
# create vectors
x = 1:4
y = 5:8
z = 9:12

rbind(x, y, z)   # row bind
cbind(x, y, z)   # column bind
matrix(c(x, y, z), nrow=3, ncol=4, byrow=TRUE)
```


### Lists

<span class="objclass">Lists</span> in R are highly flexible objects, and probably the most commonly used for data science. Unlike vectors, whose elements must be of the same type, lists can contain anything as their elements, even other lists.

Here is a list. We use the <span class="func">list</span> function to create it. 

```{r list}
x = list(1, "apple", list(3, "cat"))
x
```

We often want to loop some function over a list.

```{r listloop}
for (elem in x) print(class(elem))
```

Lists can, and often do, have named elements.

```{r namedlist}
x = list("a" = 25, "b" = -1, "c" = 0)
x["b"]
```

Almost all standard models in base R and other packages return an object that is a list.  Knowing how to work with a list will allow you to easily access the contents of the model object for further processing.


<div class='note'>
Python has similar structures, <span class="objclass">lists</span> and <span class="objclass">dictionaries</span>, where the latter works similarly to R's named list.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

### Data Frames

Data frames are a very commonly used data structure. Elements of a data frame can be different types, and this is because the <span class="objclass">data.frame</span> class is actually just a list.  As such, everything about lists applies to them. But they can also be indexed by row or column as well, just like matrices.  There are other very common types of object classes associated with packages that are both a <span class="objclass">data.frame</span> and some other type of structure (e.g. <span class="objclass">tibbles</span> in the tidyverse).

Usually your data frame will come directly from import or manipulation of other R objects (e.g. matrices).  However, you should know how to create one from scratch.


#### Creating a data frame

The following will create a data frame with two columns, `a` and `b`.

```{r createdf, eval=TRUE}
mydf = data.frame(a = c(1,5,2), 
                  b = c(3,8,1))
```

Much to the disdain of the tidyverse, we can add row names also.

```{r dfrownames, eval=TRUE}
rownames(mydf) = paste0('row', 1:3)
mydf
```

Everything about lists applies to data.frames, so we can add, select, and remove elements of a data frame just like lists. However, we'll visit this later, because we'll have much more flexibility with data frames than we would lists.


### **Data Structure Exercises**

#### Excercise #1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span> (use the <span class="func">class</span> or <span class="func">str</span> functions) on the object you just created.

```{r dsex1, echo=F}
mydf = data.frame(A=1:3, B=letters[1:3])
```


#### Exercise #2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created.

```{r dsex2, echo=F}
mylist = list(c('a', 'b'), 1:3, mydf)
```



#### Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?


[^tidynomatrix]: Despite what the tidyverse would have you believe.