## Vectorization & Apply

```{r vasetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```


#### Boolean Indexing

<span class="objclass">Logicals</span> are objects with values of `TRUE` or `FALSE`.

Assume x is a vector of numbers.

```{r boolindex, eval=T}
x = c(-1, 2, 10, -5)
idx = x > 2
idx
x[idx]
```



#### Flexiblity

We don't have to create an explicit index object first, as R indexing is ridiculously flexible.

```{r flexindex}
x[x > 2]
x[x != 3]
x[ifelse(x > 2  & x !=10, T, F)]
x[{y = idx; y}]
x[resid(lm(y ~ x)) > 0]
```



#### Vectorized operations

Consider the following unfortunately coded loop:

```{r loop}
for (i in 1:nrow(mydf)) {
  check = mydf$x[i] > 2
  if (check==TRUE) {
    mydf$y[i] = 'Yes'
  } 
  else {
    mydf$y[i] = 'No'
  }
}
```

Compare:

```{r boolnoloop}
mydf$y = 'No'
mydf$y[mydf$x > 2] = 'Yes'
```

This gets us the same thing, and would be much faster. Boolean indexing is an example of a vectorized operation.

The whole vector is considered:

- Rather than each element individually
- Any preprocessing is done once rather than n times

This is always faster.

Example: Log all values in a matrix.

```{r vecmatrixop}
mymatrix_log = log(mymatrix)
```

Way faster than looping over elements, rows or columns. Here we'll let the <span class="func">apply</span> function stand in for our loop, logging the elements of each column.

```{r loopvsvec, eval=T}
mymatrix = matrix(runif(100), 10, 10)
identical(apply(mymatrix, 2, log), log(mymatrix))

library(microbenchmark)
microbenchmark(apply(mymatrix, 2, log),
               log(mymatrix))
```


Many vectorized functions already exist in R.

They are often written in C, Fortran etc., and so even faster.
  

#### Apply functions

A family of functions allows for a succinct way of looping.

Common ones include:

- <span class="func">apply</span>
    - arrays, matrices, data.frames
- <span class="func">lapply</span>, <span class="func">sapply</span>, <span class="func">vapply</span>
    - lists, data.frames, vectors
- <span class="func">tapply</span>
    - grouped operations (table apply)
- <span class="func">mapply</span>
    - multivariate version of sapply
- <span class="func">replicate</span>
    - similar to sapply
    
#### Example

Standardizing variables.

```{r loopvsapply1, eval=FALSE}
for (i in 1:ncol(mydf)){
  x = mydf[,i]
  for (j in 1:length(x)){
    x[j] = (x[j] - mean(x))/sd(x)
  }
}
```

The above would be a really bad way to use R. The following is much cleaner and now you'd have a function you can use elsewhere.  

<span class="func">apply</span> will take a matrix, and apply a function over the margin, row or column, you want.

```{r loopvsapply2, eval=F}
stdize <- function(x) {
  (x-mean(x))/sd(x)
}

apply(mydf, 2, stdize)   # 1 for rows, 2 for columnwise application
```



#### Timings

The previous demonstrates how to use apply.

However, there is a <span class="func">scale</span> function in base R that uses a more vectorized approach under the hood.

The following demonstrates various approaches to standardizing the columns of the matrix, even using a parallelized approach. 

The base R function requires very little code and beats the others.

```{r timings, echo=T}
mymat = matrix(rnorm(100000), ncol=1000)

stdize <- function(x) {
  (x-mean(x)) / sd(x)
}

doubleloop = function() {
  for (i in 1:ncol(mymat)) {
    x = mymat[, i]
    for (j in 1:length(x)) {
      x[j] = (x[j] - mean(x)) / sd(x)
    }
  }
}


singleloop = function() {
  for (i in 1:ncol(mymat)) {
    x = mymat[, i]
    x = (x - mean(x)) / sd(x)
  }
}


library(parallel)
cl = makeCluster(8)
clusterExport(cl, c('stdize', 'mymat'))
doParallel::registerDoParallel(cl)

test = microbenchmark::microbenchmark(doubleloop=doubleloop(),
                                      singleloop=singleloop(), 
                                      apply=apply(mymat, 2, stdize), 
                                      parApply=parApply(cl, mymat, 2, stdize),
                                      vectorized=scale(mymat), times=25)
stopCluster(cl)
test
```
```{r vectorization_timings, echo=FALSE, eval=TRUE, cache=FALSE}
load('data/vectorization.RData'); library(microbenchmark); test
```

    
#### Apply functions

Benefits
  
- Cleaner/simpler code
- Environment kept clear of unnecessary objects
- Potentially more reproducible
    - more likely to use generalizable functions
- Might be faster
- Parallelizable
    
<span class="emph">NOT</span> necessarily faster than explicit loops.

- single loop over columns was as fast as apply
- Replicate and mapply are especially slow

ALWAYS can *potentially* be faster than loops.

  - Parallelization: parApply, parLapply etc.



#### Personal experience

I use R every day, and rarely use explicit loops.

- Note: no speed difference for a <span class="func">for</span> loop vs. using <span class="func">while</span>
- If you must use an explicit loop, create an empty object of the dimension/form you need, and then fill it in.
    - Notably faster

I pretty much never use an explicit double loop, as a little more thinking about the problem will usually provide a more efficient path to solving the problem.



#### Apply functions

Apply functions and similar approaches should be a part of your regular R experience.

Other versions we'll talk about have been optimized, but you need to know the basics in order to use those.

Any you still may need parallel versions.



#### **Vector/Apply Exercises**


##### Exercise 1

With the following <span class="objclass">matrix</span>, use <span class="func">apply</span> and the <span class="func">sum</span> function to get row or column sums.  

```{r vectorapp_ex1, echo=1}
x = matrix(1:9, 3, 3)
apply()
```

##### Exercise 2

With the following <span class="objclass">list</span>, use <span class="func">lapply</span> and <span class="func">sapply</span> and the <span class="func">sum</span> function to get sums for the elements.  There is no margin to specify with on a list, so just supply the list and the sum function.  

```{r vectorapp_ex2, echo=1}
x = list(1:3, 4:6, 7:9)
lapply()
sapply()
```

<span class="func">sapply</span> is actually just a wrapper for <span class="func">lapply</span>.  If you supply the argument `simplified=F`, it is identical. Otherwise, it attempts to return a vector or matrix.
