## Pipes

```{r pipessetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T)
```



Note: More detail on much of the following is given in [another workshop](https://m-clark.github.io/workshops/dplyr/mainSlides.html).



Pipes are operators that send what comes before the pipe to what comes after.

There are many different pipes, and some packages that use their own.

However, the vast majority of packages use the same pipe:


<div style="text-align:center;font-variant:small-caps; font-size:500%; color:#ff5503">%>%</div>


Here, we'll focus on their use with the <span class="pack">dplyr</span> package. Later, we'll use it for visualizations.

Example:

```{r pipes}
mydf %>% 
  select(var1, var2) %>% 
  filter(var1 == 'Yes') %>% 
  summary
```

> Start with a data.frame <span class="" style="font-family:Consolas">%>%</span> 
<br>&nbsp;&nbsp;&nbsp;  select columns from it <span class="" style="font-family:Consolas">%>%</span> 
<br>&nbsp;&nbsp;&nbsp;  filter/subset it <span class="" style="font-family:Consolas">%>%</span> 
<br>&nbsp;&nbsp;&nbsp;  get a summary


#### Using variables as they are created

One nice thing about pipelines is that we can use variables as soon as they are created, with out having to break out separate objects/steps.

```{r varfly}
mydf %>% 
  mutate(newvar1 = var1 + var2,
         newvar2 = newvar1/var3) %>% 
  summarise(newvar2avg = mean(newvar2))
```


### Pipes for Visualization (more later)

The following provides a means to think about pipes for visualization. It's just a generic example for now, but we'll see more later.

```{r pipeviz}
basegraph %>% 
  points %>%
  lines %>%
  layout
```


### The dot

Most functions are not 'pipe-aware' by default.

Example: pipe to a modeling function.

```{r pipeerror}
mydf %>% 
  lm(y ~ x)  # error
```

Other pipes could work in this situation.

  - e.g. <span class="func">%$%</span> in <span class="pack">magrittr</span>.

But generally, when you come upon this situation, you can use a dot.

- The dot refers to the object before the previous pipe.

```{r pipedot}
mydf %>% 
  lm(y ~ x, data=.)  # . == mydf
```


### Flexibility

Piping is not just for data.frames.

- The following starts with a character vector.
- Sends it to a recursive function (named <span class="func">..</span>).
    - <span class="func">..</span> is created on-the-fly, and has a single argument (`.`).
- After the function is created, it's used on ., whcih represents the string before the pipe.
- Result: pipes between the words.

```{r ridiculous, eval=T, echo=-1}
library(magrittr)
c('Ceci', "n'est", 'pas', 'une', 'pipe!') %>%
{
  .. <-  . %>%
    if (length(.) == 1)  .
    else paste(.[1], '%>%', ..(.[-1]))
  ..(.)
} 
```

> Put that in your pipe and smoke it René Magritte!


### Summary

Pipes are best used interactively.

Extremely useful for data exploration.

Common in many visualization packages.

See the <span class="pack">magrittr</span> package for more pipes.


