## Tidyverse


```{r tvsetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T)
```

### What is the tidyverse?

The tidyverse consists of a few key packages-

- <span class="pack">ggplot2</span>: data visualization
- <span class="pack">dplyr</span>: data manipulation
- <span class="pack">tidyr</span>: data tidying
- <span class="pack">readr</span>: data import
- <span class="pack">purrr</span>: functional programming, e.g. alternate approaches to apply
- <span class="pack">tibble</span>: tibbles, a modern re-imagining of data frames

And of course the <span class="pack">tidyverse</span> package which will load all of the above.

See also: <span class="pack">lubridate</span>, <span class="pack">rvest</span>, <span class="pack">stringr</span> and others in the 'hadleyverse'.

### What is tidy?

<span class="emph">Tidy data</span> refers to data arranged in a way that makes data processing, analysis, and visualization simpler.

In a tidy data set: 

- Each variable must have its own column.
- Each observation must have its own row.
- Each value must have its own cell.

Think *long* before *wide*.


### dplyr

<span class="pack">dplyr</span> provides a grammar of data manipulation (like <span class="pack">ggplot2</span> does for visualization).

It is the next iteration of <span class="pack">plyr</span>, but there is no longer any need for plyr really.

It's focused on tools for working with data frames.

- Over 100 functions

It has three main goals:

- Make the most important data manipulation tasks easier.
- Do them faster.
- Use the same interface to work with data frames, data tables or database.


Some key operations:


- <span class="func">select</span>: grab columns
    - select helpers: <span class="func">one\_of</span>, <span class="func">starts\_with</span>, <span class="func">num_range</span> etc.
- <span class="func">filter</span>/<span class="func">slice</span>: grab rows
- <span class="func">group_by</span>: grouped operations
- <span class="func">mutate</span>/<span class="func">transmute</span>: create new variables
- <span class="func">summarize</span>: summarize/aggregate
- <span class="func">do</span>: arbitrary operations

Various join/merge functions:

- <span class="func">inner_join</span>, <span class="func">left_join</span> etc.

Little things like:

- <span class="func">n</span>, <span class="func">n\_distinct</span>, <span class="func">nth</span>, <span class="func">n\_groups</span>, <span class="func">count</span>, <span class="func">recode</span>, <span class="func">between</span>

No need to quote variable names.


### An example

Let's say we want to select from our data the following variables:

  - Start with the **ID** variable
  - The variables **X1:X10**, which are not all together, and there are many more *X* columns
  - The variables **var1** and **var2**, which are the only *var* variables in the data
  - Any variable that starts with **XYZ**
    
How might we go about this?

##### Some base R approaches

Tedious, or multiple objects just to get the columns you want.

```{r baseRexample1, eval=FALSE}
# numeric indexes; not conducive to readibility or reproducibility
newData = oldData[,c(1,2,3,4, etc.)]

# explicitly by name; fine if only a handful; not pretty
newData = oldData[,c('ID','X1', 'X2', etc.)]

# two step with grep; regex difficult to read/understand
cols = c('ID', paste0('X', 1:10), 'var1', 'var2', grep(colnames(oldData), '^XYZ', value=T))
newData = oldData[,cols]

# or via subset
newData = subset(oldData, select = cols)
```


##### More
What if you also want observations where **Z** is **Yes**, Q is **No**, and only the observations with the top 50 values of **var2**, ordered by **var1** (descending)?

```{r baseRexample2, eval=FALSE}
# three operations and overwriting or creating new objects if we want clarity
newData = newData[oldData$Z == 'Yes' & oldData$Q == 'No',]
newData = newData[order(newData$var2, decreasing=T)[1:50],]
newData = newData[order(newData$var1, decreasing=T),]
```

And this is for fairly straightforward operations.


##### The dplyr way

```{r pipeExample, eval=FALSE}
newData = oldData %>% 
  filter(Z == 'Yes', Q == 'No') %>% 
  select(num_range('X', 1:10), contains('var'), starts_with('XYZ')) %>% 
  top_n(n=50, var2) %>% 
  arrange(desc(var1))
```



### An alternative


<span class="pack">dplyr</span> and piping is an *alternative*

- you can do all this sort of stuff with base R
- <span class="func">with</span>, <span class="func">within</span>, <span class="func">subset</span>, <span class="func">transform</span>, etc.

Even though the initial base R approach depicted is fairly concise, it still can potentially be: 

- noisier
- less legible
- less amenable to additional data changes
- requires esoteric knowledge (e.g. regular expressions)
- often requires creation of new objects (even if we just want to explore)
- often slower, possibly greatly


### tidyr

Two primary functions for manipulating data

- <span class="func">gather</span>: wide to long
- <span class="func">spread</span>: long to wide

Other useful functions include:

- <span class="func">unite</span>: paste together multiple columns into one
- <span class="func">separate</span>: complement of unite



### Example

```{r gather_spread, eval=T}
library(tidyr)
stocks <- data.frame( time = as.Date('2009-01-01') + 0:9,
                      X = rnorm(10, 0, 1),
                      Y = rnorm(10, 0, 2),
                      Z = rnorm(10, 0, 4) )
stocks %>% head
stocks %>% gather(stock, price, -time) %>% head
```

Note that the latter is an example of <span class="emph">tidy</span> data while the former is not.


### Personal Opinion

The <span class="pack">dplyr</span> grammar is clear for a lot of standard data processing tasks, and some not so common.

Extremely useful for data exploration and visualization.

- No need to create/overwrite existing objects
- Can overwrite columns as they are created
- Makes it easy to look at anything, and do otherwise tedious data checks

Drawbacks:

- Not as fast as <span class="pack">data.table</span> or even some base R approaches for many things
- The *mindset* can make for unnecessary complication
    - e.g. There is no need to pipe to create one new variable
    
    
### On the horizon

<span class="pack">multidplyr</span>

- Partitions the data across a cluster.
- Faster than data.table (after partitioning)



### **dplyr Exercises**

#### Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r, eval=FALSE}
install.packages('ggplot2movies')
library(ggplot2movies)
```


#### Exercise 1

Using the movies data set, perform each of the following actions separately.

##### Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```{r, eval=FALSE}
data %>% 
  mutate(newvar = '?')
```

```{r, echo=FALSE}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

##### Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r, echo=FALSE}
movies %>% 
  filter(year >= 2000) 
```

##### Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the title, year, budget, length, rating and votes variables.  There are at least 3 ways to do this.

```{r, echo=FALSE}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```



#### Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.
Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r, echo=FALSE}
movies %>% 
  group_by(year) %>% 
  summarise(AvgBudget=mean(budget, na.rm=T)) %>% 
  tail
```


#### Exercise 3

Now put it all together in one set of piped operations. 

- Filter movies released *after* 1990
- select the same variables as before but also the `mpaa`, `Action`, and `Drama` variables
- group by `mpaa` *and* `Action` *or* `Drama`
- get the average rating


```{r, echo=FALSE}
movies %>%
  filter(year>=1990) %>% 
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>% 
  group_by(mpaa, Drama) %>% 
  summarise(AvgRating = mean(rating))
```


