## data.table


Another package for data processing that has been useful to many is <span class="pack">data.table</span>. It works in a notably different way than <span class="pack">dplyr</span>. However, you'd use it for the same reasons, e.g. subset, grouping, update, ordered joins etc., but with key advantages in speed and memory efficiency. Like <span class="pack">dplyr</span>, the data objects are both <span class="objclass">data.frames</span> and a package specific class.


```{r dterror, eval=TRUE}
library(data.table)
dt = data.table(x=sample(1:10, 6), g=1:3, y=runif(6))
class(dt)
```

### Basics

In general, <span class="pack">data.table</span> works with brackets as in base R data frames.  However, in order to use <span class="pack">data.table</span> effectively you'll need to forget the data frame similarity. The brackets actually work like a function call, with several key arguments.  Consider the following notation to start.

```{r dtgeneric, eval=FALSE}
x[i, j, by, keyby, with = TRUE, ...]
```


Importantly: *you don't use the brackets as you would with data.frames*. What **i** and **j** can be are fairly complex. 

In general, you use **i** for filtering by rows.
 
```{r dtrows, results='hold'}
dt[2]
dt[2,]
```

You use **j** to select (by name!) or create new columns.  We can define a new column with the <span class="func">:=</span> operator.

 
```{r dtcols, results='hold'}
dt[,x]
dt[,z := x+y]  # dt now has a new column
dt[,z]
dt[g>1, mean(z), by=g]
dt
```

Because **j** is an argument, dropping columns is awkward.

```{r dtdropcol, eval=1:2, results='hold'}
dt[,-y]             # creates negative values of y
dt[,-'y', with=F]   # drops y, but now needs quotes
dt[,y:=NULL]        # drops y, but this is just a base R approach
dt$y = NULL
```


Data table does not make unnecessary copies.  For example if we do the following...

```{r dtmemory}
DT = data.table(A=5:1,B=letters[5:1])
DT2 = DT
```

DT2 and DT are just names for the same table.  You'd actually need to use the copy function to make an explicit copy.

```{r dtcopy}
DT2 = copy(DT)     
```


### Grouped operations

We can now attempt a 'group-by' operation, along with creation of a new variable. Note that these operations actually modify the <span class="objclass">dt</span> object *in place*, a key distinction with <span class="pack">dplyr</span>.  Fewer copies means less of a memory hit.

```{r dtgroup}
dt1 = dt2 = dt
dt[,sum(x,y), by=g]                  # sum of all x and y values
dt1[,newvar := sum(x), by=g]         # add new variable to the original data 
dt1
```



We can also create groupings on the fly. For a new summary data set, we'll take the following approach- we create a grouping based on whether g is a value of one or not, then get the mean and sum of x for those two categories.  The corresponding dplyr approach is also shown (but not evaluated) for comparison.

```{r dtgroup2, eval=1}
dt2[, list(mean_x = mean(x), sum_x = sum(x)), by = g==1]
dt2 %>% group_by(g==1) %>% summarise(mean_x=mean(x), sum_x=sum(x))
```



### Faster!



As mentioned, the reason to use <span class="pack">data.table</span> is speed.  If you have large data or large operations it'll be useful.


#### Joins

Joins can not only be faster but and easy to do. Note that the `i` argument can be a <span class="objclass">data.table</span> object itself.  I compare its speed to the comparable <span class="pack">dplyr's</span> <span class="func">left_join</span> function.

```{r dtjoin, eval=FALSE}
dt1 = setkey(dt1, x)
dt1[dt2]
```

```{r dtjoin_timing, cache=FALSE, echo=FALSE}
load('data/timing.RData')
timing_join %>% pander::pander(justify='lr')
# timing_join =
#   microbenchmark::microbenchmark(dt_join = dt1[dt2],
#                                  dplyr_join = left_join(dt1, dt2, by='x')) %>%
#   summary() %>%
#   select(expr, mean) %>%
#   mutate(mean=round(mean, 2)) %>%
#   rename(func=expr, `mean (microseconds)`=mean)
```

#### Group by

We can use the <span class="func">setkey</span> function to order a data set by a certain column(s).  This ordering is done by reference; again, no copy is made.  Doing this will allow for faster grouped operations, though you likely will only see the speed gain with very large data. 

```{r dtsetkey, echo=1:6,eval=1:8}
test_dt0 = data.table(x=rnorm(10000000), g = sample(letters, 10000000, replace=T))
test_dt1 = copy(test_dt0)
test_dt2 = setkey(test_dt1, g)

identical(test_dt0, test_dt1)
identical(test_dt1, test_dt2)

timing_group_by_1 %>% pander::pander(justify='lr')


# timing_group_by_1 = 
#   microbenchmark::microbenchmark(test_dt0 = test_dt0[,mean:=mean(x),by=g],
#                                  test_dt1 = test_dt1[,mean:=mean(x),by=g],
#                                  test_dt2 = test_dt2[,mean:=mean(x),by=g]) %>% 
#   summary() %>% 
#   select(expr, mean) %>% 
#   mutate(mean=round(mean, 2)) %>% 
#   rename(func=expr, `mean (milliseconds)`=mean)


```

#### String matching

The <span class="func">chin</span> function returns a vector of the *positions* of (first) matches of its first argument in its second, where both arguments are character vectors.  Essentially it's just like the <span class="func">%in%</span> function for character vectors. 

Consider the following. We sample the first 14 letters 1000 times with replacement and see which ones match in a subset of another subset of letters.  I compare the same operation to <span class="pack">stringr</span> and the <span class="pack">stringi</span> package whose functionality it's using.

```{r dtcharacters, echo=2:3, eval=2:4}
set.seed(1234)
lets_1 = sample(letters[1:14], 1000, replace=T)
lets_1 %chin% letters[13:26] %>% head(10)
timing_chmatch %>% pander::pander(justify='lr')

# library(stringr); library(stringi)
# timing_chmatch =
#   microbenchmark::microbenchmark(
#   dt = lets_1 %chin% letters[13:26],
#   stringi= stri_detect_regex(lets_1, paste(letters[13:26], collapse='|')),
#   stringr= str_detect(lets_1, paste(letters[13:26], collapse='|'))
# ) %>%
#   summary() %>%
#   select(expr, mean) %>%
#   mutate(mean=round(mean, 2)) %>%
#   rename(func=expr, `mean (microseconds)`=mean)


```




#### Reading files

If you use <span class="pack">data.table</span> for nothing else, you'd still want to consider it strongly for reading in large text files.  The function <span class="func">fread</span> may be quite useful in being memory efficient too.  I compare it to <span class="pack">readr</span>.

```{r dtfread, eval=FALSE}
fread('data/cars.csv')
```

```{r dtfread_timing, echo=FALSE}
# library(readr)
# timing_read =
#   microbenchmark::microbenchmark(
#   dt = fread('data/cars.csv'),
#   readr = read_csv('data/cars.csv')
# ) %>%
#   summary() %>%
#   select(expr, mean) %>%
#   mutate(mean=round(mean, 2)) %>%
#   rename(func=expr, `mean (microseconds)`=mean)

timing_read %>% pander::pander(justify='lr')
```



#### More speed

The following demonstrates some timings from [here](http://stackoverflow.com/questions/3505701/r-grouping-functions-sapply-vs-lapply-vs-apply-vs-tapply-vs-by-vs-aggrega/34167477#34167477).  I reproduced it on my own machine based on 50 million observations. The grouped operations that are applied are just a sum and length on a vector. 


By the way, never, ever use <span class="func">aggregate</span>. For anything.

```{r dttimings, eval=TRUE, echo=FALSE}
as.data.table(sapply(timing_group_by_big, `[[`, "elapsed"), keep.rownames = TRUE)[,.(fun = V1, elapsed = V2)][order(-elapsed)]
```

Ever.

Really.

### Pipe with data.table

Piping can be done with data.table objects too, using the brackets, but it's awkward at best.

```{r dtpipebad, eval=F}
mydt[,newvar:=mean(x),][,newvar2:=sum(newvar), by=group][,-'y', with=FALSE]
mydt[,newvar:=mean(x), 
  ][,newvar2:=sum(newvar), by=group
  ][,-'y', with=FALSE
  ]
```

Probably better to just use a standard pipe and dot approach if you really need it.

```{r dtpipebad2, eval=F}
mydt[,newvar:=mean(x),] %>% 
  .[,newvar2:=sum(newvar), by=group] %>% 
  .[,-'y', with=FALSE]
```

### Summary

Faster methods are great to have. If you have large data this is one package that can help.

  - Especially for group-by and joins.

Drawbacks:

- Complex
- The syntax can be awkward
- It doesn't work like a data.frame, which can be confusing
- Piping with brackets isn't really feasible
- Does not have its own 'verse'


If speed and/or memory is (potentially) a concern, <span class="pack">data.table</span>.

For interactive exploration, <span class="pack">dplyr</span>.

Piping allows one to use both, so no need to choose.

And on the horizon...

### dtplyr

Coming soon to an R near you, or possibly abandonware, only time will tell.  However, what the <span class="pack">dtplyr</span> package purports to do is implement the <span class="pack">data.table</span> back-end for <span class="pack">dplyr</span> so that you can seamlessly use them together.  The following shows times for a grouped operation of a data frame of two variables, a random uniform draw of 5e7 values, and a grouping variable of 500k groups.

```{r dtplyr, echo=FALSE, eval=TRUE, cache=F}
# something weird with n() function in chunk; saved out
# detach(package:data.table); library(dplyr); library(dtplyr)
# set.seed(123)
# n = 5e7  possible conflict with n()?
# k = 5e5
# x = runif(5e7)
# grp = sample(k, 5e7, TRUE)
# 
# dtplyrtest = system.time({
#     dt = tbl_dt(data.frame(x, grp))
#     r.dplyr = summarise(group_by(dt, grp), sum(x), n())
# })
# timingsdtplyr = append(timing[6:7], list(dtplyr=dtplyrtest))
# save(timingsdtplyr, 'data/timingsdtplyr.RData')
load('data/timingsdtplyr.RData')
data_frame(package=names(timingsdtplyr)) %>% 
  mutate(timing = sapply(timingsdtplyr, `[[`, "elapsed")) %>% 
  arrange(desc(timing)) %>% 
  pander::pander()
```

Just for giggles I did the same in Python with a <span class="pack">pandas</span> <span class="objclass">DataFrame</span> and group-by operation, and it took over seven seconds.  But remember, R is slow!

<div class='note'>
As of this writing, <span class="pack">dplyr</span> went through a notable update that changed some core functionality. Unfortunately, <span class="pack">dtplyr</span> has not been updated since then, so it will likely only work for the most standard of operations at present.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

### **data.table Exercises**


#### Exercise 0

Install and load the <span class="pack">data.table</span> package.

Create the following data table.

```
mydt = data.table(expand.grid(x=1:3, y=c('a','b','c')), z=sample(1:20, 9))
```

#### Exercise 1

Create a new object that contains only the 'a' group. Think back to how you use a logical to select rows.

```{r, echo=FALSE, eval=FALSE}
mydt_a = mydt[y=='a',]
```


#### Exercise 2

Create a new object that is the sum of z grouped by x. You don't need to name the sum variable.


```{r, echo=FALSE, eval=FALSE}
mydt_grpsum = mydt[,sum(z), by=x]
mydt_grpsum = mydt[,list(sumz=sum(z)), by=x]
```
