## **Indexing**

```{r indexsetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

What follows is a refresher.  Presumably you've had enough R exposure to be aware of some of this. However, much of data processing regards data frames, or other tables of mixed data types, so more time will be spent on slicing and dicing of data frames instead.  Even so, it would be impossible to use R effectively without knowing how to handle basic data types.

### Base R Indexing Refresher

##### Slicing vectors

```{r vectorSlice, eval=TRUE}
letters[4:6]
letters[c(13,10,3)]
```

##### Slicing matrices/data.frames

```{r matrixSlice}
myMatrix[1, 2:3]
```


##### Label-based indexing

```{r dfindexlabel}
mydf['row1', 'b']
```

##### Position-based indexing

```{r dfindexpos}
mydf[1, 2]
```


##### Mixed indexing

```{r dfindexmix}
mydf['row1', 2]
```

If the row/column value is empty, all rows/columns are retained.

```{r dfindexslice}
mydf['row1',]
mydf[,'b']
```


##### Non-contiguous

```{r dfindexnoncont}
mydf[c(1,3),]
```

##### Boolean

```{r dfindexbool}
mydf[mydf$a >= 2,]
```


### List/Data.frame extraction

<span class="func">[</span> : grab a slice of elements/columns

<span class="func">[[</span> : grab specific elements/columns

<span class="func">$</span> : grab specific elements/columns

<span class="func">@</span>: extract slot for S4 objects

```{r dflistslice}
my_list_or_df[2:4]
```


```{r dflistextract}
my_list_or_df[['name']]
```

```{r dflistextract2}
my_list_or_df$name
```

```{r dflistextract3}
my_list@name
```


### Indexing Exercises

This following is a refresher of base R indexing only.

Here is a <span class="objclass">matrix</span>, a <span class="objclass">data.frame</span> and a <span class="objclass">list</span>.

```{r ixex0}
mymatrix = matrix(rnorm(100), 10, 10)
mydf = cars
mylist = list(mymatrix, thisdf = mydf)
```

#### Exercise 1

For the <span class="objclass">matrix</span>, in separate operations, take a slice of rows, a selection of columns, and a single element.

```{r ixex1, echo=F}
mymatrix[1:5, ]
mymatrix[, 1:5]
mymatrix[1,2]
```

#### Exercise 2

For the <span class="objclass">data.frame</span>, grab a column in 3 different ways.

```{r ixex2, echo=F}
mydf$disp
mydf[,2]
mydf['disp']
```

#### Exercise 3

For the <span class="objclass">list</span> grab an element by number and by name.

```{r ixex3, echo=F}
mylist[2]
mylist$thisdf
```
