## **Pipes**

```{r pipessetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T)
```


<div class='note'>
More detail on much of the following is provided [here](https://m-clark.github.io/workshops/dplyr/mainSlides.html).

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

<br>

Pipes are operators that send what comes before the pipe to what comes after. There are many different pipes, and some packages that use their own. However, the vast majority of packages use the same pipe:


<div style="text-align:center;font-variant:small-caps; font-size:500%; color:#ff5500">%>%</div>


Here, we'll focus on their use with the <span class="pack">dplyr</span> package, and the tidyverse more generally. Pipes are also utilized heavily in visualization.

Example:

```{r pipes}
mydf %>% 
  select(var1, var2) %>% 
  filter(var1 == 'Yes') %>% 
  summary
```

```
Start with a data.frame %>% 
    select columns from it %>% 
    filter/subset it %>% 
    get a summary
```


#### Using variables as they are created

One nice thing about pipelines is that we can use variables as soon as they are created, with out having to break out separate objects/steps.

```{r varfly}
mydf %>% 
  mutate(newvar1 = var1 + var2,
         newvar2 = newvar1/var3) %>% 
  summarise(newvar2avg = mean(newvar2))
```


### Pipes for Visualization (more later)

The following provides a means to think about pipes for visualization. It's just a generic example for now, but we'll see more later.

```{r pipeviz}
basegraph %>% 
  points %>%
  lines %>%
  layout
```


### The dot

Most functions are not 'pipe-aware' by default. In the following we try to send our data frame to <span class="func">lm</span> for a regression.

```{r pipeerror}
mydf %>% 
  lm(y ~ x)  # error
```

Other pipes could work in this situation, e.g. <span class="func">%$%</span> in <span class="pack">magrittr</span>. But generally, when you come upon this situation, you can use a dot to represent the object before the pipe.

```{r pipedot}
mydf %>% 
  lm(y ~ x, data=.)  # . == mydf
```


### Flexibility

Piping is not just for <span class="objclass"></span>, <span class="objclass"></span> and <span class="objclass"></span> can be used as well, and would be the primary object for the <span class="pack"></span> family of functions we'll discuss later.

- The following starts with a character vector.
- Sends it to a recursive function (named <span class="func">..</span>).
    - <span class="func">..</span> is created on-the-fly, and has a single argument (`.`).
- After the function is created, it's used on ., which represents the string before the pipe.
- Result: pipes between the words[^collapse].

```{r ridiculous, eval=T, echo=-1}
library(magrittr)
c('Ceci', "n'est", 'pas', 'une', 'pipe!') %>%
{
  .. <-  . %>%
    if (length(.) == 1)  .
    else paste(.[1], '%>%', ..(.[-1]))
  ..(.)
} 
```

> Put that in your pipe and smoke it René Magritte!




### Summary

Pipes are best used interactively, though you can use them within functions as well, and they are extremely useful for data exploration.  Nowadays, more and more packages are being made that are 'pipe-aware', especially many visualization packages.



<div class='note'>
See the <span class="pack">magrittr</span> package for more pipes.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>



[^collapse]: That was a very complicated way to do this `paste(c('Ceci', "n'est", 'pas', 'une', 'pipe!'), collapse=' %>% ')`.