## Tidyverse


```{r tvsetup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(echo = T, eval=T, cache.rebuild = F)
```

### What is the tidyverse?

The tidyverse consists of a few key packages:

- <span class="pack">ggplot2</span>: data visualization
- <span class="pack">dplyr</span>: data manipulation
- <span class="pack">tidyr</span>: data tidying
- <span class="pack">readr</span>: data import
- <span class="pack">purrr</span>: functional programming, e.g. alternate approaches to apply
- <span class="pack">tibble</span>: tibbles, a modern re-imagining of data frames

And of course the <span class="pack">tidyverse</span> package itself, which will load all of the above in a way that will avoid naming conflicts.

```{r tv_load, eval=FALSE}
library(tidyverse)
```
```{r tv_startup_message, echo=FALSE, message=TRUE}
"Loading tidyverse: ggplot2
Loading tidyverse: tibble
Loading tidyverse: tidyr
Loading tidyverse: readr
Loading tidyverse: purrr
Loading tidyverse: dplyr
Conflicts with tidy packages -------------------------
filter(): dplyr, stats
lag():    dplyr, stats" %>% 
  message()
```



In addition, there are other packages like <span class="pack">lubridate</span>, <span class="pack">rvest</span>, <span class="pack">stringr</span> and others in the **hadleyverse** that are also greatly useful.

### What is tidy?

<span class="emph">Tidy data</span> refers to data arranged in a way that makes data processing, analysis, and visualization simpler.

In a tidy data set: 

- Each variable must have its own column.
- Each observation must have its own row.
- Each value must have its own cell.

Think *long* before *wide*.


### dplyr

<span class="pack">dplyr</span> provides a grammar of data manipulation (like <span class="pack">ggplot2</span> does for visualization).  It is the next iteration of <span class="pack">plyr</span>, but there is no longer any need for <span class="pack">plyr</span> really.  It's focused on tools for working with data frames, with over 100 functions that might be of specific use to you.

It has three main goals:

- Make the most important data manipulation tasks easier.
- Do them faster.
- Use the same interface to work with data frames, data tables or a database.


Some key operations include:


- <span class="func">select</span>: grab columns
    - select helpers: <span class="func">one\_of</span>, <span class="func">starts\_with</span>, <span class="func">num_range</span> etc.
- <span class="func">filter</span>/<span class="func">slice</span>: grab rows
- <span class="func">group_by</span>: grouped operations
- <span class="func">mutate</span>/<span class="func">transmute</span>: create new variables
- <span class="func">summarize</span>: summarize/aggregate
- <span class="func">do</span>: arbitrary operations

There are various (SQL-like) join/merge functions:

- <span class="func">inner_join</span>, <span class="func">left_join</span> etc.

And there are a lot of little things like:

- <span class="func">n</span>, <span class="func">n\_distinct</span>, <span class="func">nth</span>, <span class="func">n\_groups</span>, <span class="func">count</span>, <span class="func">recode</span>, <span class="func">between</span>

In addition, there is no need to quote variable names.


#### An example

Let's say we want to select from our data the following variables:

  - Start with the **ID** variable
  - The variables **X1** through **X10**, which are not all together, and there are many more *X* columns
  - The variables **var1** and **var2**, which are the only variables with *var* in their name
  - Any variable with a name that starts with **XYZ**
    
How might we go about this in a dataset of possibly hundreds or even thousands of columns? There are several base R approaches that we could go with, but often they will be tedious, or require multiple objects to be created just to get the columns you want.  Let's start with the worst choice.

```{r baseRexample1a, eval=FALSE}
newData = oldData[,c(1,2,3,4, etc.)]
```

Using numeric indexes, or rather <span class="emph">magic numbers</span>, is not conducive to readability or reproducibility.  If anything changes about the data columns, the numbers may no longer be applicable, and you'd have to redo the line again.

We could name the variables explicitly.

```{r baseRexample1b, eval=FALSE}
newData = oldData[,c('ID','X1', 'X2', etc.)]
```

This would be fine if there are only a handful.  But if you're trying to reduce a 1000 column data set to several dozen it's tedious, and generally not pretty regardless.

A more advanced alternative regards a two-step approach with <span class="emph">regular expressions</span>. This requires that you know something about regex (and you should), but it is difficult to read/understand by those who don't, and often by even yourself if it's more complicated.  In any case, you first will need to create an object that represents the column names first, otherwise it looks unwieldy if used within brackets or a function like <span class="func">subset</span>.

```{r baseRexample1c, eval=FALSE}
cols = c('ID', paste0('X', 1:10), 'var1', 'var2', grep(colnames(oldData), '^XYZ', value=T))

newData = oldData[,cols]

# or via subset
newData = subset(oldData, select = cols)
```


Now consider there is even more to do. What if you also want observations where **Z** is **Yes**, Q is **No**, and only the observations with the top 50 values of **var2**, ordered by **var1** (descending)?  Probably the more straightforward way in R to do so would be something like the following, where each part is broken out and we continuously write over the object as we modify it.

```{r baseRexample2, eval=FALSE}
# three operations and overwriting or creating new objects if we want clarity
newData = newData[oldData$Z == 'Yes' & oldData$Q == 'No',]
newData = newData[order(newData$var2, decreasing=T)[1:50],]
newData = newData[order(newData$var1, decreasing=T),]
```

And this is for fairly straightforward operations.


Now consider doing all of the previous in one piped operation.  The <span class="pack">dplyr</span> package will allow us to do something like the following.

```{r pipeExample, eval=FALSE}
newData = oldData %>% 
  select(num_range('X', 1:10), 
         contains('var'), 
         starts_with('XYZ')) %>% 
  filter(Z == 'Yes', 
         Q == 'No') %>% 
  top_n(n=50, var2) %>% 
  arrange(desc(var1))
```

Even if it hadn't been explained before, you might have been able to guess a little as to what was going on.  The code is fairly succinct, we don't have to keep referencing objects repeatedly, and no explicit intermediary objects are created.

<span class="pack">dplyr</span> and piping is an *alternative*. You can do all this sort of stuff with base R, for example, with functions like <span class="func">with</span>, <span class="func">within</span>, <span class="func">subset</span>, <span class="func">transform</span>, etc.  Though the initial base R approach depicted is fairly concise, in general, it can potentially be: 

- more verbose
- less legible
- less amenable to additional data changes
- requires esoteric knowledge (e.g. regular expressions)
- often requires creation of new objects (even if we just want to explore)
- often slower, possibly greatly


### Running example

The following data was scraped initially scraped from the web as follows.  It is data from the NBA basketball league for the 2017-2018 season with things like player names, points per game, field goal percentage etc.  We'll use it as an example to demonstrate various functionality found within <span class="pack">dplyr</span>.

```{r basketballDataScrape, eval=FALSE}
library(rvest)
url = "http://www.basketball-reference.com/leagues/NBA_2018_totals.html"
bball = read_html(url) %>% 
  html_nodes("#totals_stats") %>% 
  html_table %>% 
  data.frame 
save(bball, file='data/bball.RData')
```

However you can just load it into your workspace. Note that when initially downloaded, the data is all character strings.  We'll fix this later.  The following shows the data as it will eventually be.

```{r load_bball}
load('data/bball.RData')
glimpse(bball[,1:5])
```


### Selecting Columns

Often you do not need the entire data set. While this is easily handled in base R (as shown earlier), it can be more clear to use <span class="func">select</span> in <span class='pack'>dplyr</span>.  Now we won't have to create separate objects, use quotes or $, etc.

```{r select1}
bball %>% 
  select(Player, Tm, Pos) %>% 
  head
```

What if we want to drop some variables?

```{r select2}
bball %>%     
  select(-Player, -Tm, -Pos)  %>% 
  head
```

#### Helper functions

Sometimes, we have a lot of variables to select, and if they have a common naming scheme, this can be very easy.

```{r select3}
bball %>% 
  select(Player, contains("3P"), ends_with("RB")) %>% 
  arrange(desc(TRB)) %>% 
  head
```

The <span class="func">select</span> also has helper functions to make selecting columns even easier.  I probably don't even need to explain what's being done above, and this is the power of the tidyverse way.  Here is the list of <span class="emph">helper functions</span> to be aware of:

- <span class="func">starts_with</span>: starts with a prefix
- <span class="func">ends_with</span>: ends with a suffix
- <span class="func">contains</span>: contains a literal string
- <span class="func">matches</span>: matches a regular expression
- <span class="func">num_range</span>: a numerical range like x01, x02, x03.
- <span class="func">one_of</span>: variables in character vector.
- <span class="func">everything</span>: all variables.

### Filtering Rows

There are repeated header rows in this data[^headers], so we need to drop them.  This is also why everything was character string when we first imported it, because having any character strings in a column coerces the entire column to be character,  since all elements need to be of the same type.  Character string is chosen over others because anything can be converted to a string, but not everything can be a number.  

Filtering by rows requires the basic indexing knowledge we talked about before, especially Boolean indexing. In the following, `Rk`, or rank, is really just a row id, but if it equals 'Rk' instead of a number, we know we're dealing with a header row, so we'll drop it.

```{r filter0}
bball = bball %>% 
  filter(Rk != "Rk")
```

- <span class='func'>filter</span> returns rows with matching conditions.
- <span class='func'>slice</span> allows for a numeric indexing approach[^noslice].

Say we want too look at forwards (SF or PF) over the age of 35.  The following will do this, and since some players play on multiple teams, we'll want only the unique information on the variables of interest. The function <span class="func">distinct</span> allows us to do this.


```{r filter1}
bball %>% 
  filter(Age > 35, Pos == "SF" | Pos == "PF") %>% 
  distinct(Player, Pos, Age)              
```

Maybe we want just the first 10 rows...

```{r filter2}
bball %>% 
  slice(1:10)
```

We can use filtering even with variables just created. 

```{r uniteFilterArrange}
bball %>% 
  unite("posTeam", Pos, Tm) %>%         # create a new variable
  filter(posTeam == "PF_SAS") %>%       # use it for filtering
  select(Player, posTeam, Age) %>%      # use it for selection
  arrange(desc(Age))                    # descending order 
```



### Generating New Data

One of the most common data processing tasks beyond subsetting the data is generating new variables. The function <span class='func'>mutate</span> takes a vector and returns one of the same dimension.  In addition, there is <span class="func">mutate_at</span>, <span class="func">mutate_if</span>, and <span class="func">mutate_all</span> to help with specific scenarios.


To demonstrate, we'll use <span class='func'>mutate_at</span> to make appropriate columns numeric, i.e. everything except `Player`, `Pos`, and ` Tm`.  It takes two inputs, variables and functions to apply.  As there are multiple variables and (potentially) multiple functions, we use the <span class="func">vars</span> and <span class="func">funs</span> functions to denote them[^funs].

```{r mutateAt}
bball = bball %>% 
  mutate_at(vars(-Player, -Pos, -Tm), funs(as.numeric))   

glimpse(bball[,1:7])
```

Now that the data is correctly specified, the following demonstrates how we can use the standard <span class="func">mutate</span> function to create composites of existing variables.

```{r mutate}
bball = bball %>% 
  mutate(trueShooting = PTS / (2 * (FGA + (.44 * FTA))),
         effectiveFG = (FG + (.5 * X3P)) / FGA, 
         shootingDif = trueShooting - FG.)

summary(select(bball, shootingDif))  # select and others don't have to be piped to use
```


### Grouping and Summarizing Data

A very common task is to look at group-based statistics, and we can use <span class='func'>group_by</span> and <span class="func">summarize</span> to help us in this regard[^summarise].  Base R has things like <span class="func">aggregate</span>, <span class="func">by</span>, and <span class="func">tapply</span> for this, but they should not be used, as this approach is much more straightforward, flexible, and faster.

For this, I'm going to start putting together several things we've demonstrated thus far. Ultimately we'll create a variable called <span class="objclass">trueShooting</span>, which represents 'true shooting percentage', and get an average for each position.

```{r groupby}
bball %>%   
  mutate(trueShooting = PTS / (2 * (FGA + (.44 * FTA))),
         effectiveFG = (FG + (.5 * X3P)) / FGA, 
         shootingDif = trueShooting - FG.) %>%  
  select(Player, Tm, Pos, MP, trueShooting, effectiveFG, PTS) %>% 
  group_by(Pos) %>%                                                 
  summarize(meanTrueShooting = mean(trueShooting, na.rm = TRUE))    
```

We can use <span class="func">do</span> on grouped data to go further.  This function can be used to create a new <span class="emph">list-column</span> in the data, the elements of which can be *anything*, even the results of an analysis for each group.  As such, we can use <span class="pack">tidyr's</span> <span class="func">unnest</span> to get back to a standard data frame.

To demonstrate, the following will group data by position, then get the correlation between field-goal percentage and free-throw shooting percentage.  For some reason, one individual was labeled as 'PF-C', so we change them to just 'C'.  On your own, I recommend just looking at it to the do line first, and compare to this result.

```{r do}
bball %>% 
  mutate(Pos = if_else(Pos=='PF-C', 'C', Pos)) %>% 
  group_by(Pos) %>%     
  do(FgFt_Corr=cor(.$FG., .$FT., use='complete')) %>% 
  unnest(FgFt_Corr)
```

As a reminder, data frames are lists. As such, anything can go into the 'columns'. 

```{r do2}
library(nycflights13)
carriers = group_by(flights, carrier)
group_size(carriers)

mods = do(carriers, model = lm(arr_delay ~ dep_time, data = .)) # reminder that data frames are lists
mods %>% 
  summarize(rsq = summary(model)$r.squared) %>% 
  head
```

<div class='note'>
You can use <span class="func">group_by</span> on more than one variable, e.g. `group_by(var1, var2)`

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>
 


### Merging Data

Merging data is yet another very common data task, as data often comes from multiple sources. In order to do this, we need some common identifier among the sources by which to join them. The following is a list of <span class="pack">dplyr</span> join functions.

<span class="func">inner_join</span>: return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

<span class="func">left_join</span>: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

<span class="func">right_join</span>: return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

<span class="func">semi_join</span>: return all rows from x where there are matching values in y, keeping just columns from x. It differs from an inner join because an inner join will return one row of x for each matching row of y, where a semi join will never duplicate rows of x.

<span class="func">anti_join</span>: return all rows from x where there are not matching values in y, keeping just columns from x.

<span class="func">full_join</span>: return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.


Probably the most common is a left join, where we kind of have one primary data set, and are adding data from another source to it. The following is a simple demonstration.


```{r merge_demo, echo=-(1:2), message=TRUE}
band_members = data_frame(Name = c('Seth', 'Francis', 'Bubba'),
                          Band = c('Com Truise', 'Pixies', 'The New Year'))
band_instruments = data_frame(Name = c('Seth', 'Francis', 'Bubba'),
                              Instrument = c('Synthesizer', 'Guitar', 'Guitar'))

band_members
band_instruments

left_join(band_members, band_instruments)
```

Merges can get quite complex, and involve multiple data sources. In many cases you may have to do a lot of processing before getting to the merge, but <span class="pack">dplyr's</span> joins will help quite a bit.


### tidyr

The <span class="pack">tidyr</span> package can be thought of as a specialized subset of <span class="pack">dplyr's</span> functionality, as well as an update to the previous <span class="pack">reshape</span> and <span class="pack">reshape2</span> packages[^reshape]. Some of its functions for manipulating data you'll want to be familiar with are:

- <span class="func">gather</span>: wide to long
- <span class="func">spread</span>: long to wide
- <span class="func">unite</span>: paste together multiple columns into one
- <span class="func">separate</span>: complement of unite
- <span class="func">unnest</span>: unnest 'list columns'

The following example shows how we take a 'wide-form' data set, where multiple columns represent different stock prices, and turn it into two columns, one representing stock name, and one for the price.

```{r gather_spread}
library(tidyr)
stocks <- data.frame( time = as.Date('2009-01-01') + 0:9,
                      X = rnorm(10, 0, 1),
                      Y = rnorm(10, 0, 2),
                      Z = rnorm(10, 0, 4) )
stocks %>% head
stocks %>% 
  gather(stock, price, -time) %>% 
  head()
```

Note that the latter is an example of <span class="emph">tidy data</span> while the former is not.  Why do we generally prefer such data?  Precisely because the most common data operations, grouping, filtering, etc., would work notably more efficiently with such data.  *This is especially the case for visualization*.

The following demonstrates the separate function utilized for a very common data processing task- dealing with names. Here' we'll separate <span class="objclass">player</span> into first and last names based on the space.

```{r tidyrSpread}
bball %>% 
  separate(Player, into=c('firstName', 'lastName'), sep=' ') %>% 
  select(1:5) %>% 
  head
```

Note that this won't necessarily apply to every name.  So further processing may be required.



### Personal Opinion

The <span class="pack">dplyr</span> grammar is clear for a lot of standard data processing tasks, and some not so common.

Extremely useful for data exploration and visualization.

- No need to create/overwrite existing objects
- Can overwrite columns and use as they are created
- Makes it easy to look at anything, and do otherwise tedious data checks

Drawbacks:

- Not as fast as <span class="pack">data.table</span> or even some base R approaches for many things[^multidplyr]
- The *mindset* can make for unnecessary complication
    - e.g. There is no need to pipe to create a single new variable
- Some approaches are not very intuitive
- Notably less ability to work with some very common data structures (e.g. matrices)
    

All in all, if you've only been using base R approaches, the tidyverse will change your R life!  It makes all the sorts of things you do all the time easier and clearer.  Highly recommended!



### **dplyr Exercises**

#### Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r, eval=FALSE}
install.packages('ggplot2movies')
library(ggplot2movies)
data('movies')
```


#### Exercise 1

Using the movies data set, perform each of the following actions separately.

##### Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```{r ex1a_1, eval=FALSE}
data %>% 
  mutate(new_var_name = '?')
```

```{r ex1a_2, echo=FALSE, eval=FALSE}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

##### Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r ex1b, echo=FALSE, eval=FALSE}
movies %>% 
  filter(year >= 2000) 
```

##### Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the `title`, `year`, `budget`, `length`, `rating` and `votes` variables.  There are at least 3 ways to do this.

```{r ex1c, echo=FALSE, eval=FALSE}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```



#### Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.

Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r ex2, echo=FALSE, eval=FALSE}
movies %>% 
  group_by(year) %>% 
  summarise(AvgBudget=mean(budget, na.rm=T)) %>% 
  tail
```


#### Exercise 3

Now put it all together in one set of piped operations. 

- Filter movies released *after* 1990
- select the same variables as before but also the `mpaa`, `Action`, and `Drama` variables
- group by `mpaa` *and* (your choice) `Action` *or* `Drama`
- get the average rating

It should spit out something like the following:

```{r ex3, echo=FALSE}
movies %>%
  filter(year>=1990) %>% 
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>% 
  group_by(mpaa, Drama) %>% 
  summarise(AvgRating = mean(rating))
```


[^headers]: You may be thinking- 'it's 2017 and why on earth would anyone do that?!'.  Peruse most sports websites and you'll see that fundamental web design basics escape them.  See also, financial sites.

[^summarise]: As Hadley Wickham is from New Zealand, and his examples use <span class="func">summarise</span>, you'll probably see it about as much as you do the other spelling, especially since it will come up first in autocomplete.

[^multidplyr]: There is <span class="pack">multidplyr</span>, but it doesn't appear to have been updated since well before <span class="pack">dplyr</span> itself underwent major changes.

[^noslice]: If you're following good programming practices, you'll likely never use <span class="func">slice</span>.



[^funs]: For only one function, you could have just supplied `as.numeric` as the second argument.

[^reshape]: Some still use <span class="pack">reshape2</span> but there is no reason to.