
# (PART\*) Part I: Information Processing {-}


# Data Structures

<br>
<div style = 'text-align: center'>
<i class="fas fa-globe-americas fa-5x" style = 'color:#0077be'></i><span class="" style = "margin-right: 25px"></span>
<!-- <i class="far fa-arrow-alt-circle-right fa-3x"></i><span class="" style = "margin-right: 25px"></span> -->
<i class="fas fa-database fa-5x"   aria-hidden="true"></i><span class="" style = "margin-right: 25px"></span>
<!-- <i class="far fa-arrow-alt-circle-right fa-3x"></i><span class="" style = "margin-right: 25px;"></span> -->
<i class="fas fa-code fa-5x" style = 'color:#1f65b7'></i>
</div>
<br>

The goal of data science is to use data to understand the world around you.  The primary tool of data science is a programming language that can convert human intention and collected evidence to actionable results.  The tool we'll demonstrate here is R.

In order to use R to understand the world around you, you have to know the basics of how R works.  Everything in R revolves around information in the form of data, so let's start with how data exists within R. 

R has several core data structures, and we'll take a look at each.

  
- Vectors
- Factors
- Lists
- Matrices/arrays
- Data frames

The more you know about R data structures, the more you'll know how to use them, how packages use them, and you'll also better understand why things go wrong when they do, and the further you'll be able to go with your data.  Furthermore, most of these data structures are common to many programming languages (e.g. vectors, lists, matrices), so what you learn with R will often generalize to other languages as well.

<div class='note'>
R and other programming languages are used via an IDE (integrated development environment), which makes programming vastly easier through syntax highlighting, code completion, and more.  RStudio the IDE of choice for R, while Python is varied (e.g. PyCharm for software developers, Spyder for users of  Anaconda), and others like VSCode might be useful for many languages.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

## Vectors

*Vectors* form the basis of R data structures. Two main types are <span class="objclass">atomic</span> and <span class="objclass">lists</span>, but we'll talk about lists separately.

Here is an R vector. The <span class="emph">elements</span> of the vector are numeric values.

```{r vec}
x = c(1, 3, 2, 5, 4)
x
```


All elements of an atomic vector are the same <span class="emph">type</span>. Example types include:
  
- <span class="objclass">character</span>
- <span class="objclass">numeric</span> (double)
- <span class="objclass">integer</span>
- <span class="objclass">logical</span>

In addition, there are special kinds of values like <span class="objclass">NA</span> ('not available' i.e. missing), <span class="objclass">NULL</span>, <span class="objclass">NaN</span> (not a number), <span class="objclass">Inf</span> (infinite) and so forth.

<div class='note'>
You can use <span class="func">typeof</span> to examine an object's type, or use an `is` function, e.g. <span class="func">is.logical</span>, to check if an object is a specific type.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

### Character strings

When dealing with text, objects of the character class are what you'd typically be dealing with.  

```{r create_a_char, eval=T}
x = c('... Of Your Fake Dimension', 'Ephemeron', 'Dryswch', 'Isotasy', 'Memory')
class(x)
```

Not much to it, but be aware there is no real limit to what is represented as a character vector. For example, in a data frame, a special class we'll talk about later, you could have a column where each entry is one of the works of Shakespeare.


### Factors

An important type of vector is a <span class="objclass">factor</span>. Factors are used to represent categorical data structures.  Although not exactly precise, one can think of factors as integers with labels.  For example, the underlying representation of a variable for <span class="objclass">sex</span> is 1:2 with labels 'Male' and 'Female'.  They are a special class with attributes, or metadata, that contains the information about the *levels*.

```{r factor_atts}
x = factor(rep(letters[1:3], e = 10))
x
attributes(x)
```


```{r factors, eval=F, echo=FALSE}
x = factor(1:3, labels = c('q', 'V', 'what the heck?'))
x
```


The underlying representation is numeric, but it is important to remember that factors are *categorical*.  Thus, they can't be used as numbers would be, as the following demonstrates.

```{r factors2, error=TRUE}
x_num = as.numeric(x)  # convert to a numeric object
sum(x_num)
sum(x)
```

#### Strings vs. factors

The main thing to note is that factors are generally a statistical phenomenon, and are required to do statistical things with data that would otherwise be a simple character string.  If you know the relatively few levels the data can take, you'll generally want to use factors, or at least know that statistical packages and methods may require them.  In addition, factors allow you to easily overcome the silly default alphabetical ordering of category levels in some very popular visualization packages.

For other things, such as text analysis, you'll almost certainly want character strings instead, and in many cases it will be required.  It's also worth noting that a lot of base R and other behavior will coerce strings to factors.  This made a lot more sense in the early days of R, but is not really necessary these days.


<div class='note'>
Some packages to note to help you with processing strings and factors:

- <span class="pack">forcats</span>
- <span class="pack">stringr</span>
<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

### Logicals

Logical scalar/vectors are those that take on one of two values: `TRUE` or `FALSE`. They are especially useful in flagging whether to run certain parts of code, and indexing certain parts of data structures (e.g. taking rows that correspond to `r TRUE`). We'll talk about the latter usage later.

Here is a logical vector.

```{r logical}
my_logic = c(TRUE, FALSE, TRUE, FALSE, TRUE, TRUE)
```

Note also that logicals are also treated as binary 0:1, and so, for example, taking the mean will provide the proportion of `TRUE` values.

```{r logical_demo}
!my_logic
as.numeric(my_logic)
mean(my_logic)
```

### Numeric and integer

The most common type of data structure you'll deal with are integer and numeric vectors.  

```{r int_num}
ints = -3:3   # integer sequences are easily constructed with the colon operator
class(ints)


x = rnorm(5)  # 5 random values from the standard normal distribution

x

typeof(x)
class(x)

typeof(ints)
is.numeric(ints)  # also numeric!
```

The main difference between the two is that integers regard whole numbers only and are otherwise smaller in size in memory, but practically speaking you typically won't distinguish them for most of your data science needs.


### Dates

Another common data structure you'll deal with is a date variable. Typically dates require special treatment and to work as intended, but they can be stored as character strings or factors if desired.  The following shows some of the base R functionality for this.

```{r dates}
Sys.Date()

x = as.Date(c(Sys.Date(), '2020-09-01'))

x
```


In almost every case however, a package like <span class="pack" style = "">lubridate</span> will make processing them much easier.  The following shows how to strip out certain aspects of a date using it.

```{r lubridate}
library(lubridate)

month(Sys.Date())

day(Sys.Date())

wday(Sys.Date(), label = TRUE )

quarter(Sys.Date())

as_date('2000-01-01') + 100
```

In general though, dates are treated as numeric variables, with consistent (but arbitrary) starting point.  If you use these in analysis, you'll probably want to make zero a useful value (e.g. the starting date).  

```{r numeric-date}
as.numeric(Sys.Date())

as.Date(10, origin = '2000-01-01')  # 10 days after a supplied origin
```

For visualization purposes, you can typically treat date variables as is, as ordered factors, or use the values as labels, and get the desired result.



## Matrices

With multiple dimensions, we are dealing with <span class="objclass">arrays</span>. <span class="objclass">Matrices</span> are two dimensional (2-d) arrays, and extremely commonly used for scientific computing. The vectors making up a matrix *must all be of the same type*.  For example, all values in a matrix might be <span class="objclass">numeric</span>, or all <span class="objclass">character strings</span>.



### Creating a matrix

Creating a matrix can be done in a variety of ways.

```{r createMatrix}
# create vectors
x = 1:4
y = 5:8
z = 9:12

rbind(x, y, z)   # row bind

cbind(x, y, z)   # column bind

matrix(
  c(x, y, z),
  nrow = 3,
  ncol = 4,
  byrow = TRUE
)
```


## Lists

<span class="objclass">Lists</span> in R are highly flexible objects, and probably the most commonly used for applied data science. Unlike vectors, whose elements must be of the same type, lists can contain anything as their elements, even other lists.

Here is a list. We use the <span class="func">list</span> function to create it. 

```{r list}
x = list(1, "apple", list(3, "cat"))
x
```

We often want to loop some function over a list.

```{r listloop}
for (element in x) print(class(element))
```

Lists can, and often do, have named elements, which we can then extract by name.

```{r namedlist}
x = list("a" = 25, "b" = -1, "c" = 0)
x[["b"]]
```

Almost all standard models in base R and other packages return an object that is a list.  Knowing how to work with a list will allow you to easily access the contents of the model object for further processing.


<div class='note'>
Python has similar structures, <span class="objclass">lists</span> and <span class="objclass">dictionaries</span>, where the latter works similarly to R's named list.

<img class='img-note' src="img/R.ico" style="display:block; margin: 0 auto;"> 
</div>

## Data Frames

Data frames are a very commonly used data structure, and are essentially a representation of data in a table format with rows and columns. Elements of a data frame can be different types, and this is because the <span class="objclass">data.frame</span> class is actually just a list.  As such, everything about lists applies to them. But they can also be indexed by row or column as well, just like matrices.  There are other very common types of object classes associated with packages that are both a <span class="objclass">data.frame</span> and some other type of structure (e.g. <span class="objclass">tibbles</span> in the tidyverse).

Usually your data frame will come directly from import or manipulation of other R objects (e.g. matrices).  However, you should know how to create one from scratch.


### Creating a data frame

The following will create a data frame with two columns, `a` and `b`.

```{r createdf, eval=TRUE}
mydf = data.frame(
  a = c(1, 5, 2),
  b = c(3, 8, 1)
)
```

Much to the disdain of the tidyverse, we can add row names also.

```{r dfrownames, eval=TRUE}
rownames(mydf) = paste0('row', 1:3)
mydf
```

Everything about lists applies to data.frames, so we can add, select, and remove elements of a data frame just like lists. However we'll visit this more in depth later, and see that we'll have much more flexibility with data frames than we would lists for common data analysis and visualization.


## Data Structure Exercises

### Excercise 1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span> (use the <span class="func">class</span> or <span class="func">str</span> functions on the object you just created).

```{r dsex1, echo=F}
mydf = data.frame(A = 1:3, B = letters[1:3])
```


### Exercise 2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created in Exercise 1.

```{r dsex2, echo=F}
mylist = list(c('a', 'b'), 1:3, mydf)
```



### Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?



## Link to Python Data Structures Notebook

[Available on GitHub](https://github.com/m-clark/data-processing-and-visualization/blob/master/jupyter_notebooks/dataStructures.ipynb)