# Writing Functions

You can't do anything in R without using functions, but have you ever written your own?  Why would you?

- Efficiency
- Customized functionality
- Reproducibility
- Extend the work that's already been done

There are many benefits to writing your own functions, and it's actually easy to do.  Once you get the basic concept down, you'll likely find yourself using your own functions more and more.

## A starting point

Let's assume you want to calculate the mean, standard deviation, and number of missing values for a variable, called `myvar`.  We could do something like the following

```{r func_ex, eval=FALSE}
mean(myvar)
sd(myvar)
sum(is.na(myvar))
```

Now let's say you need to do it for several variables.  Here's what your custom function could look like.  It takes a single input, the variable you want information about, and returns a data frame with that info.

```{r func_ex2}
my_summary <- function(myvar) {
  data.frame(
    mean = mean(myvar, na.rm = TRUE),
    sd = sd(myvar, na.rm = TRUE),
    N_missing = sum(is.na(myvar))
    )
}
```

Note that if there are actually missing values, we need to set `na.rm = TRUE` or the <span class="func">mean</span> and <span class="func">sd</span> will return `NA`. Let's try it. I add missing values to test that aspect.

```{r func_ex3, echo=-1}
mtcars = datasets::mtcars  # to undo previous factors
my_summary(mtcars$mpg)

mtcars2 = mtcars
mtcars2$wt[c(3,10)] = NA
my_summary(mtcars2$wt)
```

Seems to work fine.  Let's add how many total observations there are.

```{r func_ex4}
my_summary <- function(myvar) {
  data.frame(
    mean = mean(myvar, na.rm = TRUE),
    sd = sd(myvar, na.rm = TRUE),
    N_total = length(myvar),
    N_missing = sum(is.na(myvar))
    )
}
```

That was easy! Let's try it.

```{r func_ex5}
my_summary(mtcars2$wt)
```

This shows that writing the first part of any function can be straightforward. Then, once in place, you can usually add functionality without too much trouble.  Eventually you could have something very complicated, but which will make sense to you because you built it from the ground up.


Now let's do it for every column!

```{r func_ex6}
mtcars2 %>% 
  map_dfr(my_summary, .id = 'variable')
```

The <span class="func">map_dfr</span> function is just like our previous usage in the [iterative programming][Iterative programming] section, just that it will create mini-data.frames then row bind them together.


## DRY

An oft-quoted mantra in programming is ***D**on't **R**epeat **Y**ourself*.  One context regards iterative programming, where we would rather write one line of code than one-hundred.  More generally though, we would like to gain efficiency where possible.  A good rule of thumb is, if you are writing the same set of code more than twice, you should write a function to do it instead.  

Consider the following example, where we want to subset the data given a set of conditions.  Given the cylinder, engine displacement, and mileage, we'll get different parts of the data.

```{r dry, eval=FALSE}
good_mileage_displ_low_cyl_4  = if_else(cyl == 4 & displ < mean(displ) & hwy > 30, 'yes', 'no')
good_mileage_displ_low_cyl_6  = if_else(cyl == 6 & displ < mean(displ) & hwy > 30, 'yes', 'no')
good_mileage_displ_low_cyl_8  = if_else(cyl == 8 & displ < mean(displ) & hwy > 30, 'yes', 'no')
good_mileage_displ_high_cyl_4 = if_else(cyl == 4 & displ > mean(displ) & hwy > 30, 'yes', 'no')
good_mileage_displ_high_cyl_6 = if_else(cyl == 6 & displ > mean(displ) & hwy > 30, 'yes', 'no')
good_mileage_displ_high_cyl_8 = if_else(cyl == 8 & displ > mean(displ) & hwy > 30, 'yes', 'no')
```

It was tedious, but that's not much code.  But now consider- what if you want to change the mpg cutoff? The mean to median? Something else?  You have to change all of it.  Let's write a function instead!

```{r mpgfunc}
good_mileage <- function(cylinder = 4,
                         mpg_cutoff = 30,
                         displ_fun = mean,
                         displ_low = TRUE,
                         cls = 'compact') {
  
  if (displ_low == TRUE) {
    result = mpg %>%
      filter(cyl == cylinder,
             displ <= displ_fun(displ),
             hwy >= mpg_cutoff,
             class == cls)
  }
  else {
    result = mpg %>%
      filter(cyl == cylinder,
             displ >= displ_fun(displ),
             hwy >= mpg_cutoff,
             class == cls)
  }
  
  result
}
```

So what's going on here? The function takes five inputs:

- cyl: Which cylinder type we want
- mpg_cutoff: The cutoff for 'good' mileage
- displ_fun: Whether the displacement to be based on the mean or something else
- displ_low: Whether we are interested in low or high displacement vehicles
- cls: the class of the vehicle (e.g. compact or suv)

Then it simply filters the data to observations that match the input criteria, and returns that result at the end.  We also put <span class="emph">default values</span> to the arguments, which can be done to your discretion.  

The core of the function uses a <span class="emph">conditional statement</span> using standard <span class="func">if</span>...<span class="func">else</span> structure.  The <span class="func">if</span> part determines whether some condition holds.  If it does, then proceed to the next step in the brackets.  If not, skip to the <span class="func">else</span> part. You may have used the <span class="func">ifelse</span> function in base R, or <span class="pack">dplyr</span>'s <span class="func">if_else</span> as above, which are a short cuts for this approach.  We can also add conditional else statements (<span class="func">else if</span>) and nest, drop the <span class="func">else</span> entirely, nest conditionals within other conditionals, etc.  Like loops, conditional statements look very similar across all programming languages.

JavaScript:

```{javascript, eval=FALSE}
if (Math.random() < 0.5) {
console.log("You got Heads!");
} else {
console.log("You got Tails!");
}
```

Python: 

```{python, eval=FALSE}
if x == 2:
  print(x)
else:
  print(x*x)
```



In any case, with our function at the ready, we can now do the things we want to as needed:

```{r mpgfunc_demo}
good_mileage(mpg_cutoff = 40)

good_mileage(cylinder = 8, mpg_cutoff = 15, displ_low = F, cls='suv')
```

Let's extend the functionality by adding a year argument (only values available are 2008 and 1999).

```{r mpgfunc_extend, echo=1:6}
good_mileage <- function(cylinder = 4,
                         mpg_cutoff = 30,
                         displ_fun = mean,
                         displ_low = TRUE,
                         cls = 'compact',
                         yr = 2008) {
  
  if (displ_low) {
    result = mpg %>%
    filter(cyl == cylinder,
           displ <= displ_fun(displ),
           hwy >= mpg_cutoff,
           class == cls,
           year == yr)
  }
  else {
    result = mpg %>%
    filter(cyl == cylinder,
           displ >= displ_fun(displ),
           hwy >= mpg_cutoff,
           class == cls,
           year == yr)
  }
  
  result
}
```

```{r mpgfunc_extend_demo}
good_mileage(cylinder = 8, 
             mpg_cutoff = 19, 
             displ_low = F, 
             cls='suv', 
             yr = 2008)
```

So we now have something that is flexible, reusable, and extensible, and it took less code than writing out the individual lines of code


## Anonymous functions

Oftentimes we just need a quick and easy function for a one-off application, especially when using apply/map functions.  For example, both the following would calculate standard deviations of columns.

```{r lambda, eval=FALSE}
apply(mtcars, 2, sd)
apply(mtcars, 2, function(x) sd(x))
```

The difference between the two is that for the latter, our function didn't have to be a named object already available.  To further illustrate this, we'll create a robust standardization function that uses the median and median absolute deviation rather than the mean and standard deviation.

```{r lambda_ex}
# some variables have a mad = 0, and so return Inf, NaN
# apply(mtcars, 2, function(x) (x - median(x))/mad(x)) %>% 
#   head()

mtcars %>%
  map_df(function(x) (x - median(x))/mad(x))
```


Even if you don't use [anonymous functions](https://en.wikipedia.org/wiki/Anonymous_function) (sometimes called lambda functions or similar), it's important to understand them, because you'll often see other people's code using them.



## Writing Functions Exercises

### Excercise 1

Write a function that takes the log of the sum of two values (i.e. just two single numbers) using the <span class="func">log</span> function.  Just remember that within a function, you can write R code just like you normally would.

```{r wf_ex1, eval=FALSE}
log_sum <- function(a, b) {
  ?
}

```


### Excercise 1b

What happens if the sum of the two numbers is negative?  You can't take a log of a negative value, so it's an error.  How might we deal with this?  Try using a conditional to provide an error message using the stop function.  Make the message whatever you want.

```{r wf_ex1b, eval=FALSE}
log_sum <- function(a, b) {
  #
  #
  #
  
  if (?) {
    stop('Your message here.')
  } 
  else {
    return(your_log_sum_object)
  }
}

```


### Exercise 2


Let's write a function that will take a numeric variable and convert it to a character string of 'positive' vs. 'negative'.  We can use `if {}... else {}` structure, <span class="func">ifelse</span>, or <span class="pack">dplyr</span>::<span class="func">if_else</span>- they all would accomplish this.  In this case, the input is a single vector of numbers, and the output will recode any negative value to 'negative' and positive values to 'positive' (or whatever you want).  Here is an example of how we would just do it as a one-off.

```{r wf_ex2, eval=FALSE}
set.seed(123)  # so you get the exact same 'random' result
x <- rnorm(10)
if_else(x < 0, "negative", "positive")
```

Now try your hand at writing a function for that.

```{r wf_ex2b, eval=FALSE}
pos_neg <- function(?) {
  ?
}
```


### Thinking Exercises

How would you write a check for Exercise 1 to make sure the inputs are numeric and positive (can't take the log of non-positive values)?  See `?stop` and `?message` for starters.



