---
title: <span style="font-size:125%; font-variant:small-caps; ">Data Processing<span style="font-size:50%; "><br></span>Visualization</span>
author: |
  <img src="../../img/Rlogo.svg" style="display:block; margin: 0 auto; width:20%"><br>
  <div class="title"><span style="font-size:125%; font-variant:small-caps; ">Michael Clark</span><br>
  <span style="font-size:75%; margin: 0 auto;">Statistician Lead</span> <br> 
  <span style="font-size:75%; margin: 0 auto;">CSCAR</span> <br> <br></div>

output: 
  revealjs::revealjs_presentation:
    theme: solarized
    highlight: pygments
    center: true
    transition: fade
    smart: true
    background_transition: slide
    css: slides_light.css
    reveal_options:
      slideNumber: true
      previewLinks: true
      viewDistance: 3
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=FALSE, cache=FALSE,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
library(magrittr); library(dplyr); library(ggplot2);library(ggplot2movies)
```


## Outline

Part 1

>- Overview of Data Structures

>- Input/Output

>- Vectorization and Apply functions

Part 2

>- Pipes, and how to use them

>- plyr, dplyr, tidyr

>- data.table

Part 3

>- Visualization with ggplot2

>- Adding Interactivity


## Part 1

```{r, child = 'dataStructures.Rmd', eval=T}
```

```{r, child = 'IO.Rmd', eval=T}
```

```{r, child = 'indexing.Rmd', eval=T}
```

```{r, child = 'VectorApp.Rmd', eval=T}
```


## Part 2


```{r, child = 'pipes.Rmd', eval=T}
```

```{r, child = 'dplyr.Rmd', eval=T}
```

```{r, child = 'datatable.Rmd', eval=T}
```

## Part 3

```{r, child = 'ggplot.Rmd', eval=T}
```

```{r, child = 'interactive.Rmd', eval=T}
```



## Summary

With the right tools, data exploration can be: 

- easier
- faster
- more efficient
- more fun!

Use them to wring your data dry of what it has to offer.

<br><br>

<div style="font-variant:small-caps; font-size:32pt;", class='emph'>Embrace a richer understanding of your data!</div>


