---
---
```{r setupIndexing, include=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=F,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
```

## 

<div style="text-align:center; font-size:200%; font-variant:small-caps;" ><span style='color:#1e90ff'>Indexing</span></div>


## Base R Indexing Refresher

Slicing vectors

```{r vectorSlice, eval=TRUE}
letters[4:6]
letters[c(13,10,3)]
```

Slicing matrices/data.frames

```{r matrixSlice, eval=FALSE}
myMatrix[1, 2:3]
```

## Base R Indexing Refresher

Label-based indexing:

```{r dfindexlabel}
mydf['row1', 'b']
```

Position-based indexing:

```{r dfindexpos}
mydf[1, 2]
```

## Base R Indexing Refresher

Mixed indexing:

```{r dfindexmix}
mydf['row1', 2]
```

If the row/column value is empty, all rows/columns are retained.

```{r dfindexslice}
mydf['row1',]
mydf[,'b']
```

## Base R Indexing Refresher

Non-contiguous:

```{r dfindexnoncont}
mydf[c(1,3),]
```

Boolean:

```{r dfindexbool}
mydf[mydf$a >=2,]
```

## Base R Indexing Refresher

List/Data.frame extraction

<span class="func">[</span> : grab a slice of elements/columns

<span class="func">[[</span> : grab specific elements/columns

<span class="func">$</span> : grab specific elements/columns

<span class="func">@</span>: extract slot for S4 objects


## List/Data.frame extraction

```{r dflistslice}
my_list_or_df[2:4]
```


```{r dflistextract}
my_list_or_df[['name']]
```

```{r dflistextract2}
my_list_or_df$name
```
