---
---

```{r setupInteractive, include=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=TRUE, cache=FALSE,
                      R.options=list(width=120), fig.width=4, fig.height=3, fig.align = 'center')
```


## 

<div style="text-align:center;font-variant:small-caps; font-size:64pt; color:#1e90ff"></div>

```{r title, echo=F, cache=F}
nodes = data.frame(id=1:2, label=c('Interactive', 'Visualization'), color='#1e90ff')
edges = data.frame(from=1, to=2, color='transparent')
library(visNetwork)
visNetwork(nodes, edges, width=800, height=800) %>% 
  visNodes(shape='text', font=list(color='#1e90ff', size=164)) %>% 
  # visPhysics(solver='barnesHut', barnesHut=list(gravitationalConstant = -50000, springConstant=.11, springLength=50)) %>% 
  visOptions(highlightNearest=F) 
# %>%
  # visHierarchicalLayout(direction='LR')
```


## Packages

<span class="pack">ggplot2</span> is the most widely used package for visualization in R.

However, it is not interactive by default.

Many packages use <span class="pack">htmlwidgets</span>, **d3** (JavaScript library) etc. to provide interactive graphics.


## Packages

General:

  - <span class="pack">plotly</span>
    - used also in Python, Matlab, Julia, can convert ggplot2 images to interactive ones.
  - <span class="pack">ggvis</span>
    - interactive successor to to ggplot though not currently actively developed
  - <span class="pack">rbokeh</span>
    - like plotly, it also has cross program support

Specific functionality:

  - <span class="pack">DT</span>
    - interactive data tables
  - <span class='pack'>leaflet</span>
      - maps with OpenStreetMap
  - <span class='pack'>dygraphs</span>
      - time series visualization
  - <span class='pack'>visNetwork</span>
      - Network visualization
  


## Piping for Visualization

One of the advantages to piping is that it's not limited to <span class="pack">dplyr</span> style data management functions.

<span class='emph'>*Any*</span> R function can be potentially piped to.

- several examples have already been shown.

This facilitates data exploration, especially visually.

- don't have to create objects
- new variables are easily created and subsequently manipulated just for vis
- data manipulation not separated from visualization


## htmlwidgets

Many newer visualization packages take advantage of piping.

<span class='pack'>htmlwidgets</span> is a package that makes it easy to create javascript visualizations.

- i.e. what you see everywhere on the web.

The packages using it typically are pipe-oriented and produce interactive plots.



## plotly example

A couple demonstrations with <span class="pack">plotly</span>.

Note the layering as with <span class="pack">ggplot2</span>.

Piping used before plotting.

```{r plotly1, cache=FALSE}
library(plotly)
midwest %>% 
  filter(inmetro==T) %>% 
  plot_ly(x=~percbelowpoverty, y=~percollege, mode='markers') 
```



## plotly example

<span class="pack">plotly</span> has <span class="emph">modes</span>, which allow for points, lines, text and combinations.

<span class="emph">Traces</span> work similar to geoms.

```{r plotly2, eval=FALSE}
library(mgcv)

mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor),
         prediction = predict(gam(mpg~s(wt, am, bs='fs'), data=mtcars))) %>% 
  arrange(wt) %>% 
  plot_ly(x=wt, y=mpg, color=amFactor, width=800, height=500, mode='markers') %>% 
  add_trace(x=wt, y=prediction, color=amFactor, alpha=.5, hover=hovertext, name='gam prediction')
```


## plotly example


```{r plotly2b, echo=FALSE, cache=FALSE}
library(mgcv)

mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor),
         prediction = predict(gam(mpg~s(wt, am, bs='fs'), data=mtcars))) %>% 
  arrange(wt) %>% 
  plot_ly(x=~wt, y=~mpg, color=~amFactor, width=800, height=500, mode='markers') %>% 
  add_trace(x=~wt, y=~prediction, color=~amFactor, alpha=.5, hover=~hovertext, name='gam prediction')
```

## ggplotly

The nice thing about <span class="pack">plotly</span> is that we can feed a <span class="objclass">ggplot</span> to it.

It would have been easier to use <span class="func">geom_smooth</span>, so let's do so.


```{r ggplotly, cache=FALSE}
gp = mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor),
         prediction = predict(gam(mpg~s(wt), data=mtcars))) %>% 
  arrange(wt) %>% 
  ggplot(aes(x=wt, y=mpg)) +
  geom_smooth() +
  geom_point(aes(color=amFactor))
ggplotly(width='auto')
```




## dygraphs

<span class="pack">dygraphs</span> is useful for time-series.

- Uses the <span class="pack">dygraphs.js</span> library

```{r dygraphdata, fig.height=3, cache=FALSE}
library(dygraphs)
data(UKLungDeaths)
cbind(ldeaths, mdeaths, fdeaths) %>% 
  dygraph(width=800) %>% 
  dyOptions(stackedGraph = TRUE, colors=RColorBrewer::brewer.pal(3, name='Dark2')) %>%
  dyRangeSelector(height = 20)
```


## visNetwork

<span class="pack">visNetwork</span> allows for network visualizations

- Uses the <span class="pack">vis.js</span> library

```{r visNetworkinitial, echo=FALSE, cache=TRUE}
set.seed(1352)
nodes = data.frame(id = 0:5,
                   label = c('Bobby', 'Janie','Timmie', 'Mary', 'Johnny', 'Billy'),
                   group = c('friend', 'frenemy','frenemy', rep('friend', 3)),
                   value = sample(10:50, 6))
edges = data.frame(from = c(0,0,0,1,1,2,2,3,3,3,4,5,5),
                   to = sample(0:5, 13, replace = T),
                   value = sample(1:10, 13, replace = T)) %>% 
  filter(from!=to)
```

```{r visNetwork, cache=FALSE}
library(visNetwork)
visNetwork(nodes, edges, height=600, width=800) %>% 
  visNodes(shape='circle', 
           font=list(), 
           scaling=list(min=10, max=50, label=list(enable=T))) %>% 
  visLegend()
```


## data table

Use the <span class="pack">DT</span> package for interactive dataframes.

```{r datatable, cache=FALSE}
library(DT)
movies %>% 
  select(1:6) %>% 
  filter(rating>9) %>% 
  slice(sample(1:nrow(.), 50)) %>% 
  datatable(rownames=F)
```


## Shiny

<span class="emph">Shiny</span> is a framework that can essentially allow you to build an interactive website.

- Provided by RStudio developers

Most of the more recently developed visualization packages will work specifically within the shiny and rmarkdown settings.


## Interactive and Visual Data Exploration

Interactivity allows for even more dimensions to be brought to a graphic.

Interactive graphics are more fun too!

- But they must serve a purpose
- Too often they are simply distraction, and detract from the data story

Just a couple visualization packages can go a very long way.


