---
title: "Exercises for Exploring the Tidyverse"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, exercise.lines=5, exercise.eval=FALSE, cache = F)
```

<!-- ## Tidyverse Exercises -->

## Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r init_dplyr, exercise=TRUE, message=FALSE}
install.packages('ggplot2movies')
library(ggplot2movies)
data('movies')
```


## Exercise 1

Using the movies data set, perform each of the following actions separately.

#### Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```{r dpex1a, exercise=TRUE}
movies %>% 
  mutate(new_var_name = '?')
```

```{r dpex1a-hint}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

#### Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r dpex1b, exercise=TRUE, exercise.eval=FALSE}
movies %>% 
  filter(?) 
```

```{r dpex1b-hint}
movies %>% 
  filter(year >= 2000) 
```

#### Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the `title`, `year`, `budget`, `length`, `rating` and `votes` variables.  There are at least 3 ways to do this.

```{r dpex1c, exercise=TRUE, exercise.eval=FALSE}
movies %>% 
  select(?) 
```

```{r dpex1c-hint}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```

#### Exercise 1d

Rename the `length` column to `length_in_min` (i.e. length in minutes).

```{r dpex1d, exercise=TRUE, exercise.eval=FALSE}
movies %>% 
  rename(?) 
```

## Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.

Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r dpex2, exercise=TRUE, exercise.eval=FALSE}
movies %>%
  group_by(?) %>%
  summarise(?) %>%
  tail()
```

```{r dpex2-hint}
movies %>%
  group_by(year) %>%
  summarise(AvgBudget=mean(budget, na.rm=T)) %>%
  tail()
```


## Exercise 3

Use <span class="func">gather</span> to create a 'tidy' data set from the following.

```{r dpex3, eval=FALSE, exercise=TRUE, exercise.eval=FALSE}
dat = tibble(id = 1:10,
                 x = rnorm(10),
                 y = rnorm(10))
```

```{r dpex3-hint, exercise.eval=FALSE}
dat %>% 
  gather(key = ?, value = ?, -id)
```



## Exercise 4

Now put several actions together in one set of piped operations. 

- Filter movies released *after* 1990
- select the same variables as before but also the `mpaa`, `Action`, and `Drama` variables
- group by `mpaa` *and* (your choice) `Action` *or* `Drama`
- get the average rating



```{r dpex4, exercise=TRUE, exercise.eval=FALSE}
movies %>%
  filter(?) %>%
  select(?) %>%
  group_by(?) %>%
  summarise(?)
```

```{r dpex4-hint}
movies %>%
  filter(year>=1990) %>%
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>%
  group_by(mpaa, Drama) %>%
  summarise(AvgRating = mean(rating))
```


It should spit out something like the following:

```{r dpex4_show, echo=FALSE, message=FALSE}
data('movies', package='ggplot2movies')
library(dplyr)
movies %>%
  filter(year>=1990) %>%
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>%
  group_by(mpaa, Drama) %>%
  summarise(AvgRating = mean(rating))
```


