---
---

## 


<div style="text-align:center; font-size:200%; font-variant:small-caps;" ><span style='color:#1e90ff'>Data Structures</span></div>

## Data structures

R has several core data structures:

- Vectors
    - Factors
- Lists
- Matrices/arrays
- Data frames


## Vectors

<span class="objclass">Vectors</span> form the basis of R data structures.

Two main types- <span class="objclass">atomic</span> and <span class="objclass">lists</span>, but I will treat lists separately.

Here is an R <span class="objclass">vector</span>. 

The *elements* of the vector are numeric values.

```{r vec, eval=TRUE}
x = c(1, 3, 2, 5, 4)
x
```



## Vectors

All elements of an atomic vector are the same type. 

Examples include:

- characters
- numeric (double)
- integer
- logical



## Factors

A important type of vector is a <span class="objclass">factor</span>. 

Factors are used to represent categorical data structures.

```{r factors, eval=TRUE}
x = factor(1:3, labels=c('q', 'V', 'what the heck?'))
x
```



## Factors

The underlying representation is numeric.

But, factors are *categorical*.

They can't be used as numbers would be.

```{r factors2, eval=TRUE, error=TRUE}
as.numeric(x)
sum(x)
```



## Matrices

With multiple dimensions, we are dealing with <span class="objclass">arrays.</span>

<span class="objclass">Matrices</span> are 2-d arrays, and extremely commonly used.

The vectors making up a matrix *must all be of the same type*.

- e.g. all values in a matrix must be numeric.



## Creating a matrix

Creating a matrix can be done in a variety of ways.

```{r createMatrix}
# create vectors
x = 1:4
y = 5:8
z = 9:12

rbind(x, y, z)   # row bind
cbind(x, y, z)   # column bind
matrix(c(x, y, z), nrow=3, ncol=4, byrow=TRUE)
```


## Lists

<span class="objclass">Lists</span> in R are highly flexible objects.

They can contain anything as their elements, even other lists.

- unlike vectors, whose elements must be of the same type. 

Here is a list. We use the list function to create one. 

```{r list, eval=TRUE}
x = list(1, "apple", list(3, "cat"))
x
```



## Lists

We often want to loop some function over a list.

```{r listloop, eval=TRUE}
for(elem in x) class(elem)
```

Lists can, and often do, have named elements.

```{r namedlist, eval=TRUE}
x = list("a" = 25, "b" = -1, "c" = 0)
x["b"]
```



## Data Frames

<span class="objclass">data.frames</span> are a very commonly used data structure. 

They do not have to have the same type of element.  

This is because the data.frame class is actually just a list.  

- As such, everything about lists applies to them

But they can also be indexed by row or column as well.

- like matrices



## Creating a data frame

```{r createdf, eval=TRUE}
mydf = data.frame(a = c(1,5,2),
                  b = c(3,8,1))
```

We can add row names also.

```{r dfrownames, eval=TRUE}
rownames(mydf) = paste0('row', 1:3)
mydf
```
