```{r setupPipes, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=T, eval=FALSE, cache=FALSE,
                      R.options=list(width=120), fig.width=8, fig.align = 'center')
```


## 

<div style="text-align:center;font-variant:small-caps; font-size:200%; color:#1e90ff">Pipes</div>


##

Note: 

More detail on much of this part is given in another workshop.



## Pipes

Operators that send what comes before to what comes after.

There are many different pipes.

There are many packages that use their own.

However, the vast majority of packages use the same pipe:



## {.fadeslideslow}

<div style="text-align:center;font-variant:small-caps; font-size:500%; color:#ff5503"> %>% </div>



## Pipes

Here, we'll focus on their use with the dplyr package.

Later, we'll use it for visualizations.

Example:

```{r}
mydf %>% 
  select(var1, var2) %>% 
  filter(var1 == 'Yes') %>% 
  summary
```

Start with a data.frame %>%  

&nbsp;&nbsp;&nbsp;&nbsp;select columns from it %>% 
  
&nbsp;&nbsp;&nbsp;&nbsp;filter/subset it %>% 
  
&nbsp;&nbsp;&nbsp;&nbsp;get a summary

## Using variables as they are created

We can use variables as soon as they are created.

```{r}
mydf %>% 
  mutate(newvar1 = var1 + var2,
         newvar2 = newvar1/var3) %>% 
  summarise(newvar2avg = mean(newvar2))
```


## Pipes for Visualization (more later)

Generic example:

```{r}
basegraph %>% 
  points %>%
  lines %>%
  layout
```

## The dot

Most functions are not 'pipe-aware' by default.

Example: pipe to a modeling function.

```{r}
mydf %>% 
  lm(y ~ x)  # error
```

Other pipes can would work in this situation.

  - e.g. <span class="func">%$%</span> in magrittr

But generally, one can use a dot.

- The dot refers to the object before the pipe.

```{r}
mydf %>% 
  lm(y ~ x, data=.)
```


## Flexibility

Piping is not just for data.frames.

>- The following starts with a character vector.
>- Sends it to a recursive function (named <span class="func">..</span>).
>- <span class="func">..</span> is created on-the-fly. 
>- After the function is created, it's used on ., representing the string.
>- Result: pipes between the words.

```{r provocation1, eval=T, echo=-1}
library(magrittr)
c('Ceci', "n'est", 'pas', 'une', 'pipe!') %>%
{
  .. <-  . %>%
    if (length(.) == 1)  .
    else paste(.[1], '%>%', ..(.[-1]))
  ..(.)
} 
```

>- Put that in your pipe and smoke it René Magritte!

## Pipes

Pipes are best used interactively.

Extremely useful for data exploration.

Common in many visualization packages.

See the <span class="pack">magrittr</span> package for more pipes.


