---
title: "Thinking Visually"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, exercise.lines=5, exercise.eval=FALSE, cache = F)
```

# Thinking Visually Exercises



## Exercise 1

The following uses the <span class="objclass">diamonds</span> data set that comes with <span class="pack">ggplot2</span>.  Use the <span class="func">scale_color_viridis</span> function to add a more accessible palette.

```{r thinkingvis_ex1, exercise=TRUE, echo=T, eval=FALSE, cache=FALSE, out.width='50%'}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  ????
```

```{r thinkingvis_ex1-hint, eval=F, out.width='50%'}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  scale_color_viridis()
```


## Exercise 2

Now color it by the `cut`.  To use the scale_color_viridis approach, you'll have to change one of the arguments to the function (set `discrete=T`).  Reproduce this but using one of the other viridis palettes.  See the helpfile via `?scale_color_viridis` to see how to change the palette.

```{r thinkingvis_ex2, exercise=TRUE, exercise.eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(????)) +
  scale_color_viridis(discrete=T)
```



```{r thinkingvis_ex2-hint, eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=cut)) +
  scale_color_viridis(discrete=T)
```
