---
title: "Thinking Visually"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr); library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, error=FALSE, comment=NA, R.options=list(width=220),
                      exercise.lines=5, exercise.eval=FALSE, cache = FALSE, out.width='50%')

```

## Thinking Visually Exercises



### Exercise 1

The following uses the <span class="objclass">diamonds</span> data set that comes with <span class="pack">ggplot2</span>.  Use the <span class="func">scale_color_viridis</span> function to add a more accessible palette.

```{r thinkingvis_ex1, exercise=TRUE, eval=FALSE, cache=FALSE}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  ????
```

```{r thinkingvis_ex1-hint, eval=F}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  scale_color_viridis()
```


### Exercise 2

Now color it by the `cut`.  To use the scale_color_viridis approach, you'll have to change one of the arguments to the function (set `discrete=T`).  Reproduce this but using one of the other viridis palettes.  See the helpfile via `?scale_color_viridis` to see how to change the palette.

```{r thinkingvis_ex2, exercise=TRUE, exercise.eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(????)) +
  scale_color_viridis(discrete=T)
```



```{r thinkingvis_ex2-hint, eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=cut)) +
  scale_color_viridis(discrete=T)
```


## ggplot2


### Exercise 0

Load the <span class="pack">ggplot2</span> package if you haven't already, or better yet, <span class="pack">tidyverse</span>.

### Exercise 1

Create two plots, one a scatterplot (e.g. with <span class="func">geom_point</span>) and one with lines (e.g. <span class="func">geom_line</span>) with a data set of your choosing (all of the following are base R or available after loading <span class="pack">ggplot2</span>.  Some suggestions:

- <span class="objclass">faithful</span>: Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
- <span class="objclass">msleep</span>:  mammals sleep dataset with sleep times and weights etc.
- <span class="objclass">diamonds</span>: used in the slides
- <span class="objclass">economics</span>: US economic time series.
- <span class="objclass">txhousing</span>: Housing sales in TX.
- <span class="objclass">midwest</span>: Midwest demographics.
- <span class="objclass">mpg</span>: Fuel economy data from 1999 and 2008 for 38 popular models of car

Recall the basic form for ggplot.

```
ggplot(aes(x=*, y=*, other), data=*) +
  geom_*() +
  otherLayers, theme etc.
```

Themes to play with:

- <span class="func">theme_bw</span>
- <span class="func">theme_classic</span>
- <span class="func">theme_dark</span>
- <span class="func">theme_gray</span>
- <span class="func">theme_light</span>
- <span class="func">theme_linedraw</span>
- <span class="func">theme_minimal</span>
- <span class="func">theme_trueMinimal</span> (requires the visibly package and an appreciation of the Lamborghini background from the previous visualization)

```{r ggplot_ex1, exercise=TRUE, echo=FALSE, eval=FALSE}
library(ggplot2)
ggplot(aes(x=waiting, y=eruptions), data=faithful) +
  geom_point(???)
```


### Exercise 2

Play around and change the arguments to the following. You'll need to install the <span class="pack">maps</span> package. 

- For example, do points for all county midpoints. For that you'd need to change the x and y for the point geom to an aesthetic based on the longitude and latitude, as well as add its data argument to use the <span class="objclass">seats</span> data frame. 
- Make the color of the points or text based on `subregion`. This will require adding the fill argument to the polygon geom and removing the NA setting. In addition, add the argument show.legend=F (outside the aesthetic), or you'll have a problematic legend (recall what we said before about too many colors!).  Try making color based on subregion too.
- See if you can use <span class="func">element_blank</span> on a theme argument to remove the axis information.  See `?theme` for ideas.

```{r ggplot_ex2, exercise=TRUE, exercise.eval=TRUE, exercise.lines=20}
library(maps)
mi = map_data("county", "michigan")
seats = mi %>% 
  group_by(subregion) %>% 
  summarise_at(vars(lat, long), function(x) median(range(x)))

# inspect the data
# head(mi)
# head(seats)

ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = subregion), fill = NA, colour = "grey60") +
  geom_text(aes(label = subregion), data = seats, size = 1, angle = 45) +
  geom_point(x=-83.748333, y=42.281389, color='#1e90ff', size=3) +
  theme_minimal() +
  theme(panel.grid=element_blank())
```



```{r ggplot_ex2-hint, echo=FALSE, exercise.lines=20}
ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = subregion, fill = subregion, alpha= - lat), colour = "grey60", show.legend = F) +
  geom_text(aes(label = subregion), data = seats, size = 1, angle = 45) +
  geom_point(aes(x=long, y=lat), color='#1e90ff', size=3, data = seats) +
  theme_minimal() +
  theme(panel.grid=element_blank())
```