---
title: "Thinking Visually"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr); library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, error=FALSE, comment=NA, R.options=list(width=220),
                      exercise.lines=5, exercise.eval=FALSE, cache = FALSE, out.width='50%')

```

## Thinking Visually Exercises



### Exercise 1

The following uses the <span class="objclass">diamonds</span> data set that comes with <span class="pack">ggplot2</span>.  Use the <span class="func">scale_color_viridis</span> function to add a more accessible palette.

```{r thinkingvis_ex1, exercise=TRUE, eval=FALSE, cache=FALSE}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  ????
```

```{r thinkingvis_ex1-hint, eval=F}
library(ggplot2); library(viridis)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=price)) +
  scale_color_viridis()
```


### Exercise 2

Now color it by the `cut`.  To use the scale_color_viridis approach, you'll have to change one of the arguments to the function (set `discrete=T`).  Reproduce this but using one of the other viridis palettes.  See the helpfile via `?scale_color_viridis` to see how to change the palette.

```{r thinkingvis_ex2, exercise=TRUE, exercise.eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(????)) +
  scale_color_viridis(discrete=T)
```



```{r thinkingvis_ex2-hint, eval=FALSE, out.width='50%'}
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(aes(color=cut)) +
  scale_color_viridis(discrete=T)
```


## ggplot2


### Exercise 0

Load the <span class="pack">ggplot2</span> package if you haven't already, or better yet, <span class="pack">tidyverse</span>.

### Exercise 1

Create two plots, one a scatterplot (e.g. with <span class="func">geom_point</span>) and one with lines (e.g. <span class="func">geom_line</span>) with a data set of your choosing (all of the following are base R or available after loading <span class="pack">ggplot2</span>.  Some suggestions:

- <span class="objclass">faithful</span>: Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
- <span class="objclass">msleep</span>:  mammals sleep dataset with sleep times and weights etc.
- <span class="objclass">diamonds</span>: used in the slides
- <span class="objclass">economics</span>: US economic time series.
- <span class="objclass">txhousing</span>: Housing sales in TX.
- <span class="objclass">midwest</span>: Midwest demographics.
- <span class="objclass">mpg</span>: Fuel economy data from 1999 and 2008 for 38 popular models of car

Recall the basic form for ggplot.

```
ggplot(aes(x=*, y=*, other), data=*) +
  geom_*() +
  otherLayers, theme etc.
```

Themes to play with:

- <span class="func">theme_bw</span>
- <span class="func">theme_classic</span>
- <span class="func">theme_dark</span>
- <span class="func">theme_gray</span>
- <span class="func">theme_light</span>
- <span class="func">theme_linedraw</span>
- <span class="func">theme_minimal</span>
- <span class="func">theme_trueMinimal</span> (requires the visibly package and an appreciation of the Lamborghini background from the previous visualization)

```{r ggplot_ex1, exercise=TRUE, echo=FALSE, eval=FALSE}
library(ggplot2)
ggplot(aes(x=waiting, y=eruptions), data=faithful) +
  geom_point(???)
```


### Exercise 2

Play around and change the arguments to the following. You'll need to install the <span class="pack">maps</span> package. 

- For example, do points for all county midpoints. For that you'd need to change the x and y for the point geom to an aesthetic based on the longitude and latitude, as well as add its data argument to use the <span class="objclass">seats</span> data frame. 
- Make the color of the points or text based on `subregion`. This will require adding the fill argument to the polygon geom and removing the NA setting. In addition, add the argument show.legend=F (outside the aesthetic), or you'll have a problematic legend (recall what we said before about too many colors!).  Try making color based on subregion too.
- See if you can use <span class="func">element_blank</span> on a theme argument to remove the axis information.  See `?theme` for ideas.

```{r ggplot_ex2, exercise=TRUE, exercise.eval=TRUE, exercise.lines=20}
library(maps)
mi = map_data("county", "michigan")
seats = mi %>% 
  group_by(subregion) %>% 
  summarise_at(vars(lat, long), function(x) median(range(x)))

# inspect the data
# head(mi)
# head(seats)

ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = subregion), fill = NA, colour = "grey60") +
  geom_text(aes(label = subregion), data = seats, size = 1, angle = 45) +
  geom_point(x=-83.748333, y=42.281389, color='#1e90ff', size=3) +
  theme_minimal() +
  theme(panel.grid=element_blank())
```



```{r ggplot_ex2-hint, echo=FALSE, exercise.lines=20}
ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = subregion, fill = subregion, alpha= - lat), colour = "grey60", show.legend = F) +
  geom_text(aes(label = subregion), data = seats, size = 1, angle = 45) +
  geom_point(aes(x=long, y=lat), color='#1e90ff', size=3, data = seats) +
  theme_minimal() +
  theme(panel.grid=element_blank())
```

## Interactive Visualization 
### Exercise 0

Install and load the <span class="pack">plotly</span> package. Load the <span class="pack">tidyverse</span> package if necessary (so you can use <span class="pack">dplyr</span> and <span class="pack">ggplot2</span>), and install/load the <span class="pack">ggplot2movies</span> for the IMDB data.

```{r setup_IV, include=FALSE}
library(learnr); library(tidyverse)
knitr::opts_chunk$set(exercise.lines=10, out.width='75%')

```


```{r interactive_ex0, exercise=TRUE, exercise.eval=TRUE}
library(plotly)
library(ggplot2movies)
library(tidyverse)
```



### Exercise 1

Using <span class="pack">dplyr</span>, group by year, and summarize to create a new variable that is the Average rating.   Refer to the [tidyverse][tidyverse] section if you need a refresher on what's being done here. Then create a plot with <span class="pack">plotly</span> for a line or scatter plot (for the latter, use the <span class="func">add_markers</span> function). It will take the following form, but you'll need to supply the plotly arguments.

```{r interactive_ex1, exercise=TRUE}
library(ggplot2movies)

movies %>% 
  group_by(year) %>% 
  summarise(Avg_Rating=mean(rating))
  plot_ly() %>% 
  add_markers()
```

```{r interactive_ex1-hint, echo=FALSE}
library(ggplot2movies)

movies %>% 
  group_by(year) %>% 
  summarise(Avg_Rating=mean(rating)) %>% 
  plot_ly() %>% 
  add_markers(x=~year, y=~Avg_Rating)
```


### Exercise 2

This time group by year *and* Drama. In the <span class="func">summarize</span> create average rating again, but also a variable representing the average number of votes. In your plotly line, use the size and color arguments to represent whether the average number of votes and whether it was drama or not respectively. Use <span class="func">add_markers</span>.  Note that Drama will be treated as numeric since it's a 0-1 indicator. This won't affect the plot, but if you want, you might use <span class="func">mutate</span> to change it to a factor with labels 'Drama' and 'Other'.

```{r interactive_ex2, exercise=TRUE, echo=FALSE}
movies %>% 
  group_by(year, Drama) %>% 
  summarise(Avg_Rating=mean(rating),
            Avg_Votes = mean(votes)) %>% 
  plot_ly() %>% 
  add_markers()
```

```{r interactive_ex2-hint, echo=FALSE}
movies %>% 
  group_by(year, Drama) %>% 
  summarise(Avg_Rating=mean(rating),
            Avg_Votes = mean(votes)) %>% 
  plot_ly() %>% 
  add_markers(x=~year, 
              y=~Avg_Rating, 
              size=~Avg_Votes, 
              color=~Drama)
```


### Exercise 3

Create a ggplot of your own design and then use <span class="func">ggplotly</span> to make it interactive.