# Interactive Visualization

```{r title, echo=F, cache=FALSE, eval=F}
saveWidget(vn, file='vntitle.html')
library(visNetwork)
nodes = data.frame(id=1:2, label=c('Interactive', 'Visualization'), color='#1e90ff')
edges = data.frame(from=1, to=2, color='transparent')
library(visNetwork)
vn = visNetwork(nodes, edges, height=400, width=600) %>% 
  visNodes(shape='text', font=list(color='#1e90ff', size=164)) %>% 
  visOptions(highlightNearest=F) 
```

<iframe src='../img/vntitle.html', width=800, height=300, scrolling="no", frameBorder="0"></iframe>


### Packages

<span class="pack">ggplot2</span> is the most widely used package for visualization in R.

However, it is not interactive by default.

Many packages use <span class="pack">htmlwidgets</span>, **d3** (JavaScript library) etc. to provide interactive graphics.

General:

- <span class="pack">plotly</span>
  - used also in Python, Matlab, Julia, can convert ggplot2 images to interactive ones.
- <span class="pack">highcharter</span>
  - also very general wrapper for highcharts.js and works with some R packages out of the box
- <span class="pack">rbokeh</span>
  - like plotly, it also has cross program support

Specific functionality:

  - <span class="pack">DT</span>
    - interactive data tables
  - <span class='pack'>leaflet</span>
      - maps with OpenStreetMap
  - <span class='pack'>visNetwork</span>
      - Network visualization
  

### Piping for Visualization

One of the advantages to piping is that it's not limited to <span class="pack">dplyr</span> style data management functions.

<span class='emph'>*Any*</span> R function can be potentially piped to.

- several examples have already been shown.

This facilitates data exploration, especially visually.

- don't have to create objects
- new variables are easily created and subsequently manipulated just for vis
- data manipulation not separated from visualization


### htmlwidgets

Many newer visualization packages take advantage of piping.

<span class='pack'>htmlwidgets</span> is a package that makes it easy to create javascript visualizations.

- i.e. what you see everywhere on the web.

The packages using it typically are pipe-oriented and produce interactive plots.



### plotly example

A couple demonstrations with <span class="pack">plotly</span>.

Note the layering as with <span class="pack">ggplot2</span>.

Piping used before plotting.

```{r plotly1, cache=FALSE, eval=TRUE}
library(plotly)
midwest %>% 
  filter(inmetro==T) %>% 
  plot_ly(x=~percbelowpoverty, y=~percollege, mode='markers') 
```


<span class="pack">plotly</span> has <span class="emph">modes</span>, which allow for points, lines, text and combinations.

<span class="emph">Traces</span> work similar to geoms.

```{r plotly2, echo=TRUE, cache=FALSE, eval=TRUE}
library(mgcv); library(modelr)

mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste('weight:',wt, '<br>', 'mgp:', mpg, '<br>', amFactor)) %>% 
  add_predictions(gam(mpg~s(wt, am, bs='fs'), data=mtcars)) %>% 
  arrange(wt) %>% 
  plot_ly() %>%
  add_markers(x=~wt, y=~mpg, color=~amFactor, text=~hovertext, hoverinfo='text') %>% 
  add_lines(x=~wt, y=~pred, color=~amFactor, alpha=.5, name='gam prediction', showlegend=F)
```

### ggplotly

The nice thing about <span class="pack">plotly</span> is that we can feed a <span class="objclass">ggplot</span> to it.

It would have been easy to use <span class="func">geom_smooth</span> to get a similar result, so let's do so.


```{r ggplotly, cache=FALSE, eval=TRUE}
gp = mtcars %>% 
  mutate(amFactor = factor(am, labels=c('auto', 'manual')),
         hovertext = paste(wt, mpg, amFactor)) %>% 
  arrange(wt) %>% 
  ggplot(aes(x=wt, y=mpg, color=amFactor)) +
  geom_smooth(se=F) +
  geom_point(aes(color=amFactor))
ggplotly()
```




### Highcharter

<span class="pack">Highcharter</span> is also fairly useful for a wide variety of plots.

- Uses the <span class="pack">highcharts.js</span> library

```{r dygraphdata, cache=FALSE, eval=F, echo=FALSE}
# library(dygraphs)
# data(UKLungDeaths)
# cbind(ldeaths, mdeaths, fdeaths) %>% 
#   dygraph() %>% 
#   dyOptions(stackedGraph = TRUE, colors=RColorBrewer::brewer.pal(3, name='Dark2')) %>%
#   dyRangeSelector(height = 20)
```

```{r highcharts, cache=FALSE, eval=T}
library(highcharter); library("quantmod")

x = getSymbols("GOOG", auto.assign = FALSE)

highchart(width=1000) %>%
   hc_add_series_ohlc(x)
```

### visNetwork

<span class="pack">visNetwork</span> allows for network visualizations

- Uses the <span class="pack">vis.js</span> library

```{r visNetworkinitial, echo=FALSE, cache=TRUE, eval=T}
set.seed(1352)
nodes = data.frame(id = 0:5,
                   label = c('Bobby', 'Janie','Timmie', 'Mary', 'Johnny', 'Billy'),
                   group = c('friend', 'frenemy','frenemy', rep('friend', 3)),
                   value = sample(10:50, 6))
edges = data.frame(from = c(0,0,0,1,1,2,2,3,3,3,4,5,5),
                   to = sample(0:5, 13, replace = T),
                   value = sample(1:10, 13, replace = T)) %>% 
  filter(from!=to)
```

```{r visNetwork, cache=FALSE, eval=F, echo=-1}
saveWidget(vn, file='visnetwork.html')
library(visNetwork)
visNetwork(nodes, edges) %>% #, height=600, width=800
  visNodes(shape='circle', 
           font=list(), 
           scaling=list(min=10, max=50, label=list(enable=T))) %>% 
  visLegend()
```

<iframe src='../img/visnetwork.html', width=1000, height=600, scrolling="no", frameBorder="0"></iframe>


### data table

Use the <span class="pack">DT</span> package for interactive data frames.

```{r datatable, cache=FALSE, eval=TRUE}
library(DT)
ggplot2movies::movies %>% 
  select(1:6) %>% 
  filter(rating>8, !is.na(budget), votes > 1000) %>% 
  datatable()
```


### Shiny

<span class="emph">Shiny</span> is a framework that can essentially allow you to build an interactive website.

- Provided by RStudio developers

Most of the more recently developed visualization packages will work specifically within the shiny and rmarkdown settings.


### Interactive and Visual Data Exploration

Interactivity allows for even more dimensions to be brought to a graphic.

Interactive graphics are more fun too!

- But they must serve a purpose
- Too often they are simply distraction, and detract from the data story

Just a couple visualization packages can go a very long way.


### **Interactive Visualization Exercises**


#### Exercise 0

Install and load the <span class="pack">plotly</span> package. Load the <span class="pack">dplyr</span> and <span class="pack">ggplot2</span> packages if necessary.



#### Exercise 1

Using <span class="pack">dplyr</span> group by year, and summarize to create a new variable that is the Average rating.  Then create a plot with <span class="pack">plotly</span> for a line or scatter plot (for the latter, use the <span class="func">add_markers</span> function). It will take the following form:

```
data %>% 
  group_by() %>% 
  summarize() %>% 
  plot_ly() %>% 
  add_markers()
```

```{r, echo=FALSE}
movies %>% 
  group_by(year) %>% 
  summarise(Avg_Rating=mean(rating)) %>% 
  plot_ly() %>% 
  add_markers(x=~year, y=~Avg_Rating)
```


#### Exercise 2

This time group by year and Drama. In the <span class="func">summarize</span> create average rating again, but also a variable representing the average number of votes. In your plotly line, use the size and color arguments to represent whether the average number of votes and whether it was drama or not respectively. Use <span class="func">add_markers.</span>

```{r, echo=FALSE}
movies %>% 
  group_by(year, Drama) %>% 
  summarise(Avg_Rating=mean(rating),
            Avg_Votes = mean(votes)) %>% 
  plot_ly() %>% 
  add_markers(x=~year, y=~Avg_Rating, size=~Avg_Votes, color=~Drama, mode='markers')
```


#### Exercise 3

Create a ggplot of your design and then use <span class="func">ggplotly</span> to make it interactive.

