---
title: "Data Structures Exercises"
output: 
  html_notebook:
    css: notebooks.css
    highlight: pygments
    toc: yes
    toc_float: yes
    code: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', comment=NA, autodep=T)
```

## Excercise #1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span>.

```{r}
mydf = data.frame(A=1:3, B=letters[1:3])
```


## Exercise #2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created.

```{r}
mylist = list(c('a', 'b'), 1:3, mydf)
```



## Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?