---
title: "dplyr Exercises"
output: 
  html_notebook:
    css: notebooks.css
    highlight: pygments
    toc: yes
    toc_float: yes
    code: hide
---

# Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r}
install.packages('ggplot2movies')
library(ggplot2movies)
```


# Exercise 1

Using the movies data set, perform each of the following actions separately.

## Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```
data %>% 
  mutate(newvar = ?)
```

```{r}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

## Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r}
movies %>% 
  filter(year >= 2000) 
```

## Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the title, year, budget, length, rating and votes variables.  There are at least 3 ways to do this.

```{r}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```



# Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.
Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r}
movies %>% 
  group_by(year) %>% 
  summarise(AvgBudget=mean(budget, na.rm=T)) %>% 
  tail
```




# Exercise 3

Now put it all together in one set of piped operations. 

- Filter to post 1990 movies
- select the same variables as before but also the mpaa, Action, and Drama variables
- group by mpaa *and* Action *or* Drama
- get the average rating


```{r}
movies %>%
  filter(year>=1990) %>% 
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>% 
  group_by(mpaa, Drama) %>% 
  summarise(AvgRating = mean(rating))
```

