---
title: "Exercises"
description: |
  Data Processing and Visualization
output: 
  html_document:
    theme: paper
    highlight: pygments
    # code_folding: hide
    css: [../css/standard_html.css]
    toc: true
    toc_depth: 2
    toc_float: true
  # distill::distill_article:
  #   code_folding: hide  
  #   css: [../css/radix.css]
  #   toc: true
  #   toc_depth: 2
---

```{r setup, include=FALSE, eval=TRUE, cache=FALSE}
knitr::opts_chunk$set(
  echo = T,
  message = F,
  warning = F,
  comment = NA,
  autodep = T,
  eval = FALSE,
  cache.rebuild = F,
  cache = F,
  R.options = list(width = 120),
  fig.width = 8,
  fig.align = 'center'
)
```

Note that in each exercise, the code is hidden by default, allowing you to test your skills first.  On some occasions you need the code to get started, and that will be noted.  Otherwise, disclose the code if you get stuck, but try on your own or work together first. Make sure you load the packages you need for the exercise.

# Intro

## A NOTE ABOUT INSTALLING R PACKAGES ON THE LAB MACHINES

You do not have write access to the default R library folder. When you attempt to install the first package, a <span class="emph">**pop-under**</span> window will come up (look at your task bar), and ask you if you want to try someplace else.  Just click yes, as it will have selected an appropriate location.




# Data Processing

## Data Structures Exercise

### Excercise #1

Create an object that is a <span class="objclass">matrix</span> and/or a <span class="objclass">data.frame</span>, and inspect its <span class="emph">class</span> or <span class="emph">structure</span>.

```{r dsex1}
mydf = data.frame(A = 1:3, B = letters[1:3])
```

### Exercise 2

Create a <span class="objclass">list</span> of 3 elements, the first of which contains character strings, the second numbers, and the third, the <span class="objclass">data.frame</span> or <span class="objclass">matrix</span> you just created.

```{r dsex2}
mylist = list(c('a', 'b'), 1:3, mydf)
```

### Thinking Exercises

- How is a <span class="objclass">factor</span> different from a character vector?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">matrix</span>?

- How is a <span class="objclass">data.frame</span> the same as and different from a <span class="objclass">list</span>?



## I/O Exercises

### Exercise 1

Use <span class="">readr</span> and <span class="">haven</span> to read the following files. Use the url just like you would any filename.  The latter is a Stata file.  You can use the RStudio's menu approach to import the file if you want.


```{r ioex1}
'https://raw.githubusercontent.com/m-clark/data-processing-and-visualization/master/data/cars.csv'
'https://raw.githubusercontent.com/m-clark/data-processing-and-visualization/master/data/presvote.dta'

```

### Thinking Exercises

Why might you use <span class="func">read_csv</span> from the <span class="pack">readr</span> package rather than <span class="func">read.csv</span> in base R?

What is your definition of 'big' data?


## Indexing Exercises

This following is a refresher of base R indexing only.

Here is a <span class="objclass">matrix</span>, a <span class="objclass">data.frame</span> and a <span class="objclass">list</span> (show the code).

```{r}
mymatrix = matrix(rnorm(100), 10, 10)
mydf = cars
mylist = list(mymatrix, thisdf = mydf)
```

### Exercise 1

For the <span class="objclass">matrix</span>, in separate operations, take a slice of rows, a selection of columns, and a single element.

```{r}
mymatrix[1:5, ]
mymatrix[, 1:5]
mymatrix[1,2]
```

### Exercise 2

For the <span class="objclass">data.frame</span>, grab a column in 3 different ways.

```{r}
mydf$dist
mydf[,2]
mydf['dist']
```

### Exercise 3

For the <span class="objclass">list</span> grab an element by number and by name.

```{r}
mylist[2]
mylist$thisdf
```








## tidyverse Exercises

### Exercise 0

Install and load the <span class="pack">dplyr</span> <span class="pack">ggplot2movies</span> packages. Look at the help file for the `movies` data set, which contains data from IMDB.


```{r, eval=FALSE}
install.packages('ggplot2movies')
library(ggplot2movies)
data('movies')
```


### Exercise 1

Using the movies data set, perform each of the following actions separately.

#### Exercise 1a

Use <span class="func">mutate</span> to create a centered version of the rating variable.  A centered variable is one whose mean has been subtracted from it. The process will take the following form:

```{r ex1a_1, eval=FALSE}
data %>% 
  mutate(new_var_name = '?')
```

```{r ex1a_2, echo=FALSE, eval=FALSE}
movies %>% 
  mutate(ratingCen = rating - mean(rating)) 
```

#### Exercise 1b

Use <span class="func">filter</span> to create a new data frame that has only movies from the years 2000 and beyond. Use the greater than or equal operator `>=`.

```{r ex1b, echo=FALSE, eval=FALSE}
movies %>% 
  filter(year >= 2000) 
```

#### Exercise 1c

Use <span class="func">select</span> to create a new data frame that only has the `title`, `year`, `budget`, `length`, `rating` and `votes` variables.  There are at least 3 ways to do this.

```{r ex1c, echo=FALSE, eval=FALSE}
movies %>% 
  select(title, year, budget, length, rating, votes) 
movies %>% 
  select(1:6)
movies %>% 
  select(-num_range('r',1:10), -mpaa, -starts_with('A'), -Comedy, -starts_with('D'), -Romance, -Short)
```


#### Exercise 1d

Rename the `length` column to `length_in_min` (i.e. length in minutes).




### Exercise 2

Use <span class="func">group_by</span> to group the data by year, and <span class="func">summarize</span> to create a new variable that is the average budget. The <span class="func">summarize</span> function works just like <span class="func">mutate</span> in this case.

Use the <span class="func">mean</span> function to get the average, but you'll also need to use the argument `na.rm = TRUE` within it because the earliest years have no budget recorded.


```{r ex2, echo=FALSE, eval=FALSE}
movies %>% 
  group_by(year) %>% 
  summarise(AvgBudget=mean(budget, na.rm=T)) %>% 
  tail
```


### Exercise 3

Use <span class="func">pivot_longer</span> to create a 'tidy' data set from the following.  Try to create a data frame of three columns of `id`, `group`, and `value`.

```{r ex3, eval=FALSE, echo=1:3}
N = 100

dat = tibble(
  id = 1:N,
  a = rnorm(N),
  b = rpois(N, lambda = 5),
  c = runif(N),
  d = rbinom(N, size = 1, prob = .5)
)
```

```{r ex3_hint, eval=FALSE, echo=FALSE}
dat %>% 
  pivot_longer(cols = ?,
               names_to = ?,
               values_to = ?)
```




### Exercise 4

Now put several actions together in one set of piped operations. 

- Filter movies released *after* 1990
- select the same variables as before but also the `mpaa`, `Action`, and `Drama` variables
- group by `mpaa` *and* (your choice) `Action` *or* `Drama`
- get the average rating

It should spit out something like the following:

```{r ex4, echo=FALSE}
movies %>%
  filter(year>=1990) %>% 
  select(title, year, budget, length, rating, votes, mpaa, Action, Drama) %>% 
  group_by(mpaa, Drama) %>% 
  summarise(AvgRating = mean(rating))
```

## data.table Exercises


### Exercise 0

Install and load the <span class="pack">data.table</span> package.

Create the following data table.

```
mydt = data.table(expand.grid(x=1:3, y=c('a','b','c')), z=sample(1:20, 9))
```

### Exercise 1

Create a new object that contains only the 'a' group. Think back to how you use a logical to select rows.

```{r}
mydt_a = mydt[y=='a',]
```


### Exercise 2

Create a new object that is the sum of z grouped by x. You don't need to name the sum variable.


```{r}
mydt_grpsum = mydt[,sum(z), by=x]
mydt_grpsum = mydt[,list(sumz=sum(z)), by=x]
```




# Programming


## Objects Exercises

### Exercise 1

With one function, find out what the class, number of rows, number of columns are of the following object, including what kind of object the last three columns are.

```{r tibble_inspect, eval=FALSE}
library(dplyr)
?starwars
```


## Iterative Programming Exercises

### Exercise 1

With the following <span class="objclass">matrix</span>, use <span class="func">apply</span> and the <span class="func">sum</span> function to get row or column sums of the matrix <span class="objclass">x</span>.

```{r app_ex1, echo=1, eval=FALSE}
x = matrix(1:9, 3, 3)
apply()
```

### Exercise 2

With the following list object <span class="objclass">x</span>, use <span class="func">lapply</span> and <span class="func">sapply</span> and the <span class="func">sum</span> function to get sums for the elements.  There is no margin to specify for a list, so just supply the list and the sum function.  

```{r app_ex2, echo=1, eval=FALSE}
x = list(1:3, 4:10, 11:100)
lapply()
sapply()
```


### Exercise 3

As in the previous example, use a <span class="func" style = "">map</span> function to create a data frame of the column means. See `?map` to see all your options.

```{r purrr_ex3, echo=1:5, eval=FALSE}
d = data_frame(
  x = rnorm(100),
  y = rnorm(100, 10, 2),
  z = rnorm(100, 50, 10),
)

d %>% map_df(mean)
```


## Writing Functions Exercises

### Excercise 1

Write a function that takes the log of the sum of two values (i.e. just two single numbers) using the <span class="func">log</span> function.  Just remember that within a function, you can write R code just like you normally would.

```{r wf_ex1, eval=FALSE}
log_sum <- function(a, b) {
  ?
}
```


### Excercise 1b

What happens if the sum of the two numbers is negative?  You can't take a log of a negative value, so it's an error.  How might we deal with this?  Try using a conditional to provide an error message using the <span class="func">stop</span> function.  The first part is basically identical to the function you just did.  But given that result, you will need to check for whether it is negative or not.  The message can be whatever you want.

```{r wf_ex1b, eval=FALSE}
log_sum <- function(a, b) {
  
  ?
  
  if (? < 0) {
    stop('Your message here.')
  } else {
    ?
    return(your_log_sum_results)    # this is an arbitrary name, change accordingly
  }
}
```


### Exercise 2


Let's write a function that will take a numeric variable and convert it to a character string of 'positive' vs. 'negative'.  We can use `if {}... else {}` structure, <span class="func">ifelse</span>, or <span class="pack">dplyr</span>::<span class="func">if_else</span>- they all would accomplish this.  In this case, the input is a single vector of numbers, and the output will recode any negative value to 'negative' and positive values to 'positive' (or whatever you want).  Here is an example of how we would just do it as a one-off.

```{r wf_ex2, eval=FALSE}
set.seed(123)  # so you get the exact same 'random' result
x <- rnorm(10)
if_else(x < 0, "negative", "positive")
```

Now try your hand at writing a function for that.

```{r wf_ex2b, eval=FALSE}
pos_neg <- function(?) {
  ?
}
```

## More
### Code Style Exercises

#### Exercise 1 

For the following model related output, come up with a name for each object.

```{r name_ex, eval=FALSE}
lm(hwy ~ cyl, data = mpg)                 # hwy mileage predicted by number of cylinders

summary(lm(hwy ~ cyl, data = mpg))        # the summary of that

lm(hwy ~ cyl + displ + year, data = mpg)  # an extension of that
```


#### Exercise 2


Fix this code.

```{r fix_code, eval=FALSE}
x=rnorm(100, 10, 2)
y=.2* x+ rnorm(100)
data = data.frame(x,y)
q = lm(y~x, data=data)
summary(q)
```

### Vectorization Exercises

Before we do this, did you remember to fix the names in the previous exercise?

#### Exercise 1

Show a non-vectorized (e.g. a loop) and a vectorized way to add a two to the numbers 1 through 3.

```{r vector_ex1, eval=FALSE}
?
```


#### Exercise 2

Of the following <span class="objclass">matrix</span>, which do you think is faster?  Test it.

```{r vector_ex2, eval=FALSE}
x = matrix(rpois(100000, lambda = 5), ncol = 100)
colSums(x)
apply(x, 2, sum)

microbenchmark::microbenchmark(
  cs = colSums(x),
  app = apply(x, 2, sum)
)
```

### Regex Exercises

#### Exercise 1

Using stringr and str_replace, replace all the states a's with nothing.

```{r regex_ex, eval=FALSE}
library(stringr)
str_replace(state.name, pattern = ?, replacement = ?)
```



# Modeling

## Modeling Exploration Exercises

### Exercise 1

With the Google app data, use a standard linear model (i.e. <span class="func" style = "">lm</span>) to predict one of three target variables of your choosing:

- `rating`: the user ratings of the app
- `avg_sentiment_polarity`: the average sentiment score (positive vs. negative) for the app
- `avg_sentiment_subjectivity`: the average subjectivity score (subjective vs. objective) for the app

For prediction use the following variables:

- `reviews`: number of reviews
- `type`: free vs. paid
- `size_in_MB`: size of the app in megabytes

I would suggest preprocessing the number of reviews- dividing by 100,000, scaling (standardizing), or logging it (for the latter you can add 1 first to deal with zeros[^plusonelog]).

Interpret the results.  Visualize the difference in means between free and paid apps. See the [emmeans][visualization] example above.

```{r ex1-goog, eval=FALSE}
load('data/google_apps.RData')

model = lm(? ~ reviews + type + size_in_MB, data = google_apps)

plot(emmeans::emmeans(model, ~type))
```


### Exercise 2

Rerun the above with interactions of the number of reviews or app size (or both) with type (via `a + b + a:b` or just `a*b` for two predictors).  Visualize the interaction.  Does it look like the effect differs by type?

```{r ex2-goog, eval=FALSE}
model = lm(? ~ reviews + type*?, data = google_apps)

plot(ggeffects::ggpredict(model, terms = c('size_in_MB', 'type')))
```

##  Model Criticism Exercises

#### Exercise 0

Recall the [google app exercises][model exploration exercises], we use a standard linear model (i.e. <span class="func" style = "">lm</span>) to predict one of three target variables:

- `rating`: the user ratings of the app
- `avg_sentiment_polarity`: the average sentiment score (positive vs. negative) for the app
- `avg_sentiment_subjectivity`: the average subjectivity score (subjective vs. objective) for the app

For prediction use the following variables:

- `reviews`: number of reviews
- `type`: free vs. paid
- `size_in_MB`: size of the app in megabytes

After that we did a model with an interaction.

Either using those models, or running new ones with a different target variable, conduct the following exercises.

```{r ex0-goog, eval=FALSE}
load('data/google_apps.RData')
```

### Exercise 1

Assess the model fit and performance of your first model. Perform additional diagnostics to assess how the model is doing (e.g. plot the model to look at residuals).

```{r ex1-model-assess, eval=FALSE}
summary(model)
plot(model)
```


### Exercise 2

Compare the model with the interaction model.  Based on AIC or some other metric, which one would you choose?  Visualize the interaction model if it's the better model.

```{r ex2-model-compare, eval=FALSE}
anova(model1, model2)
AIC(model1, model2)
```




## Machine Learning Exercises

### Exercise 1

Use the <span class="pack" style = "">ranger</span> package to predict the google variable `rating` by several covariates.  Feel free to just use the standard function approach rather than all the tidymodels stuff if you want, but do use a training and test approach. You can then try the model again with a different tuning.  For the first go around, 

```{r ex1-ml, eval=FALSE}
# run these if needed to load data and install the package
# load('data/google_apps.RData')
# install.packages('ranger')

google_for_mod = google_apps %>% 
  select(avg_sentiment_polarity, rating, type,installs, reviews, size_in_MB, category) %>% 
  drop_na()

google_split = google_for_mod %>% 
  initial_split(prop = 0.75)

google_train = training(google_split)
google_test  = testing(google_split)

ga_rf_results = rand_forest(mode = 'regression', mtry = 2, trees = 1000) %>%
  set_engine(engine = "ranger") %>% 
  fit(
    rating ~ ?,
    google_train
  )

test_predictions = predict(ga_rf_results, new_data = google_test)

rmse = yardstick::rmse(
  data = bind_cols(test_predictions, google_test),
  truth = rating,
  estimate = .pred
)

rsq = yardstick::rsq(
  data = bind_cols(test_predictions, google_test),
  truth = rating,
  estimate = .pred
)

bind_rows(
  rmse,
  rsq
)
```


### Exercise 2

Respecify the neural net model demonstrated above as follows, and tune over the number of hidden units to have.

```{r ml-ex-2, eval=FALSE}
grid_search = expand.grid(
  hidden_units = c(25, 50),  
  penalty = exp(seq(-4, -.25, length.out = 5))
)

happy_nn_spec = mlp(mode = 'regression',
                    penalty = tune(),
                    hidden_units = tune()) %>% 
  set_engine(engine = "nnet")

nn_tune = tune_grid(
  happy_prepped,                  # from previous examples, see tuning for regularized regression
  model = happy_nn_spec,
  resamples = happy_folds,        # from previous examples, see tuning for regularized regression
  grid = grid_search
)

show_best(nn_tune, metric = "rmse", maximize = FALSE, n = 1) 
```



# Visualization

## Thinking Visually Exercises

### Exercise 1

The following uses the <span class="objclass">diamonds</span> data set that comes with <span class="pack">ggplot2</span>.  Use the <span class="func">scale_color_viridis</span> or <span class="func">scale_color_scico</span> function to add a more accessible palette. Use `?` to examine your options.

```{r thinkingvis_ex1, echo=T, eval=FALSE, cache=FALSE}
# devtools::install_github("thomasp85/scico") # to use scientific colors
library(ggplot2)

ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(aes(color = price)) +
  ????
```

```{r thinkingvis_ex1b, eval=F, out.width='50%'}
library(ggplot2)

ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(aes(color = price), alpha = .05) +
  scale_color_scico(palette = 'acton')
```



### Exercise 2

Now color it by the `cut` instead of `price`.  Use <span class="func">scale_color_viridis/scioc_d</span>.   See the helpfile via `?scale_color_*` to see how to change the palette.

```{r thinkingvis_ex2, eval=FALSE, out.width='50%'}
ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(aes(color = cut)) +
  scale_color_scico_d(palette = 'batlow')
```


### Thinking exercises

For your upcoming presentation, *who* is your audience?


## ggplot2 Exercises

### Exercise 0

Load the <span class="pack">ggplot2</span> package if you haven’t already.

### Exercise 1

Create two plots, one a scatterplot (e.g. with <span class="func">geom_point</span>) and one with lines (e.g. <span class="func">geom_line</span>) with a data set of your choosing (all are base R or available after loading <span class="pack">ggplot2</span>.  Some suggestions:

- <span class="objclass">faithful</span>: Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
- <span class="objclass">msleep</span>:  mammals sleep dataset with sleep times and weights etc.
- <span class="objclass">diamonds</span>: used in the slides
- <span class="objclass">economics</span>: US economic time series.
- <span class="objclass">txhousing</span>: Housing sales in TX.
- <span class="objclass">midwest</span>: Midwest demographics.
- <span class="objclass">mpg</span>: Fuel economy data from 1999 and 2008 for 38 popular models of car

Recall the basic form for ggplot.

```
ggplot(aes(x=*, y=*, other), data=*) +
  geom_*() +
  otherLayers, theme etc.
```

Themes to play with:

- <span class="func">theme_bw</span>
- <span class="func">theme_classic</span>
- <span class="func">theme_dark</span>
- <span class="func">theme_gray</span>
- <span class="func">theme_light</span>
- <span class="func">theme_linedraw</span>
- <span class="func">theme_minimal</span>

```{r gg-ex1}
library(ggplot2)

ggplot(aes(x = waiting, y = eruptions), data = faithful) +
  geom_point()
```


### Exercise 2

Play around and change the arguments to the following.

```{r gg-ex2}
library(maps)

mi = map_data("county", "michigan")

seats = mi %>%
  group_by(subregion) %>%
  summarise_at(vars(lat, long), function(x)
    median(range(x)))

ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = group), fill = NA, colour = "grey60") +
  geom_text(
    aes(label = subregion),
    data = seats,
    size = 1,
    angle = 45
  ) +
  geom_point(
    y = 42.281389,
    x = -83.748333,
    color = '#1e90ff',
    size = 3
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```


## Interactive Visualization Exercises


### Exercise 0

Install and load the <span class="pack">plotly</span> package. Load the <span class="pack">dplyr</span> and <span class="pack">ggplot2</span> packages if necessary.



### Exercise 1


Using <span class="pack">dplyr</span> group by year, and summarize to create a new variable that is the Average rating.  Then create a plot with <span class="pack">plotly</span> for a line or scatter plot (for the latter, use the <span class="func">add_markers</span> function). It will take the following form:

```
data %>% 
  group_by() %>% 
  summarize() %>% 
  plot_ly() %>% 
  add_markers()
```

```{r plotly-ex1, echo=FALSE}
movies %>%
  group_by(year) %>%
  summarise(Avg_Rating = mean(rating)) %>%
  plot_ly() %>%
  add_markers(x =  ~ year, y =  ~ Avg_Rating)
```


### Exercise 2

This time group by year and Drama. In the <span class="func">summarize</span> create average rating again, but also a variable representing the average number of votes. In your plotly line, use the size and color arguments to represent whether the average number of votes and whether it was drama or not respectively. Use <span class="func">add_markers.</span>

```{r plotly-ex2, echo=FALSE}
movies %>%
  group_by(year, Drama) %>%
  summarise(Avg_Rating = mean(rating),
            Avg_Votes = mean(votes)) %>%
  plot_ly() %>%
  add_markers(
    x =  ~ year,
    y =  ~ Avg_Rating,
    size =  ~ Avg_Votes,
    color =  ~ Drama,
    mode = 'markers'
  )
```


### Exercise 3

Create a ggplot of your design and then use <span class="func">ggplotly</span> to make it interactive.

