---
title: "R II: Programming"
subtitle: 'Exercises'
output: 
  learnr::tutorial:
    ace_theme: chrome
css: css/standard_html.css
highlight: pygments
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr); library(tidyverse)
knitr::opts_chunk$set(
  echo = FALSE,
  # eval = FALSE,    # this will keep code from running regardless of chunk specific option
  exercise.lines = 5,
  exercise.eval = FALSE,
  cache = FALSE
)
```

# Programming

## Objects Exercises

### Exercise 1

With one function, find out what the class, number of rows, number of columns are of the following object, including what kind of object the last three columns are.

```{r obj_ex1, exercise = TRUE}
library(dplyr)
starwars
```

```{r obj_ex1-hint}
str()
```



## Iterative Programming Exercises

### Exercise 1

With the following <span class="objclass">matrix</span>, use <span class="func">apply</span> and the <span class="func">sum</span> function to get row or column sums of the matrix <span class="objclass">x</span>.

```{r app_ex1, exercise = TRUE}
x = matrix(1:9, 3, 3)
```

```{r app_ex1-hint}
apply()
```

### Exercise 2

With the following list object <span class="objclass">x</span>, use <span class="func">lapply</span> and <span class="func">sapply</span> and the <span class="func">sum</span> function to get sums for the elements.  There is no margin to specify for a list, so just supply the list and the sum function.  

```{r app_ex2, exercise = TRUE}
x = list(1:3, 4:10, 11:100)
```

```{r app_ex2-hint}
lapply()
sapply()
```

### Exercise 3

As in the previous example, use a <span class="func" style = "">map</span> function to create a data frame of the column means. See `?map` to see all your options.

```{r purrr_ex3, exercise = TRUE}
d = tibble(
  x = rnorm(100),
  y = rnorm(100, 10, 2),
  z = rnorm(100, 50, 10),
)
```


```{r purrr_ex3-hint}
d %>% map_df(some_function)
```



## Writing Functions Exercises

### Exercise 1

Write a function that takes the log of the sum of two values (i.e. just two single numbers) using the <span class="func">log</span> function.  Just remember that within a function, you can write R code just like you normally would.

```{r wf_ex1, exercise = TRUE}
log_sum <- function(a, b) {
  ?
}
```


### Exercise 1b

What happens if the sum of the two numbers is negative?  You can't take a log of a negative value, so it's an error.  How might we deal with this?  Try using a conditional to provide an error message using the <span class="func">stop</span> function.  The first part is basically identical to the function you just did.  But given that result, you will need to check for whether it is negative or not.  The message can be whatever you want.

```{r wf_ex1b, exercise = TRUE, exercise.lines=15}
log_sum <- function(a, b) {
  
  ?
  
  if (? < 0) {
    stop('Your message here.')
  } 
  else {
    ?
    return(your_log_sum_results)    # this is an arbitrary name, change accordingly
  }
}
```


### Exercise 2


Let's write a function that will take a numeric variable and convert it to a character string of 'positive' vs. 'negative'.  We can use `if {}... else {}` structure, <span class="func">ifelse</span>, or <span class="pack">dplyr</span>::<span class="func">if_else</span>- they all would accomplish this.  In this case, the input is a single vector of numbers, and the output will recode any negative value to 'negative' and positive values to 'positive' (or whatever you want).  Here is an example of how we would just do it as a one-off.

```{r wf_ex2, eval=FALSE}
set.seed(123)  # so you get the exact same 'random' result
x <- rnorm(10)
if_else(x < 0, "negative", "positive")
```

Now try your hand at writing a function for that.

```{r wf_ex2b, exercise = TRUE}
pos_neg <- function(?) {
  ?
}
```



## More

### Coding Style Exercises
#### Exercise 1 

For the following model related output, come up with a name for each object.

```{r name_ex, exercise = TRUE}
lm(hwy ~ cyl, data = mpg)                 # hwy mileage predicted by number of cylinders

summary(lm(hwy ~ cyl, data = mpg))        # the summary of that

lm(hwy ~ cyl + displ + year, data = mpg)  # an extension of that
```


#### Exercise 2


Fix this code.

```{r fix_code, exercise = TRUE}
x=rnorm(100, 10, 2)
y=.2* x+ rnorm(100)
data = data.frame(x,y)
q = lm(y~x, data=data)
summary(q)
```

### Vectorization Exercises

Before we do this, did you remember to fix the names in the previous exercise?

#### Exercise 1

Show a non-vectorized (e.g. a loop) and a vectorized way to add a two to the numbers 1 through 3.

```{r vector_ex1, exercise = TRUE}
?
```


#### Exercise 2

Of the following <span class="objclass">matrix</span>, which do you think is faster?  Test it.

```{r vector_ex2, exercise = TRUE}
x = matrix(rpois(100000, lambda = 5), ncol = 100)

vec = colSums(x)
loop = apply(x, 2, sum)

identical(vec, loop)

bench::mark(
  cs = colSums(x),
  app = apply(x, 2, sum),
  time_unit = 'ms'
)
```

### Regex Exercises

#### Exercise 1

Using stringr and str_replace, replace all the states a's with nothing.

```{r regex_ex, exercise = TRUE}
library(stringr)

str_replace(state.name, pattern = ?, replacement = ?)
```