---
title: "ggplot2 Exercises"
output: 
  html_notebook:
    css: notebooks.css
    highlight: pygments
    toc: yes
    toc_float: yes
    code: hide
---

# Exercise 0

Install and load the <span class="pack">ggplot2</span> package.

# Exercise 1

Create two plots, one a scatterplot (e.g. with <span class="func">geom_point</span>) and one with lines (e.g. <span class="func">geom_line</span>) with a data set of your choosing (all are base R or available after loading <span class="pack">ggplot2</span>.  Some suggestions:

- <span class="objclass">faithful</span>: Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
- <span class="objclass">msleep</span>:  mammals sleep dataset with sleep times and weights etc.
- <span class="objclass">diamonds</span>: used in the slides
- <span class="objclass">economics</span>: US economic time series.
- <span class="objclass">txhousing</span>: Housing sales in TX.
- <span class="objclass">midwest</span>: Midwest demographics.
- <span class="objclass">mpg</span>: Fuel economy data from 1999 and 2008 for 38 popular models of car

Recall the basic form for ggplot.

```
ggplot(aes(x=*, y=*, other), data=*) +
  geom_*() +
  otherLayers, theme etc.
```

Themes to play with:

- <span class="func">theme_bw</span>
- <span class="func">theme_classic</span>
- <span class="func">theme_dark</span>
- <span class="func">theme_gray</span>
- <span class="func">theme_light</span>
- <span class="func">theme_linedraw</span>
- <span class="func">theme_minimal</span>

```{r}
library(ggplot2)
ggplot(aes(x=waiting, y=eruptions), data=faithful) +
  geom_point()
```


## Exercise 2

Unhide the code and play around with it.

```{r}
library(maps)
mi <- map_data("county", "michigan")
mid_range <- function(x) mean(range(x))
seats <- plyr::ddply(mi, "subregion", plyr::colwise(mid_range, c("lat", "long")))
ggplot(mi, aes(long, lat)) +
  geom_polygon(aes(group = group), fill = NA, colour = "grey60") +
  geom_text(aes(label = subregion), data = seats, size = 1, angle = 45) +
  geom_point(y=42.281389, x=-83.748333, color='#1e90ff', size=3) +
  theme_minimal() +
  theme(panel.grid=element_blank())
```

