---
title: "Interactive Exercises"
output: 
  html_notebook:
    css: notebooks.css
    highlight: pygments
    toc: yes
    toc_float: yes
    code: hide
---

# Exercise 0

Install and load the <span class="pack">plotly</span> package. Load the <span class="pack">dplyr</span> and <span class="pack">ggplot2</span> packages if necessary.



# Exercise 1

Using <span class="pack">dplyr</span> group by year, and summarize to create a new variable that is the Average rating.  Then create a plot with <span class="pack">plotly</span> for a line or scatter plot (for the latter, use the argument `mode='markers'`). It will take the following form:

```
data %>% 
  group_by() %>% 
  summarize() %>% 
  plot_ly()
```

```{r}
sapply(c('dplyr', 'ggplot2movies', 'plotly'), library, character.only=T)

movies %>% 
  group_by(year) %>% 
  summarise(Avg_Rating=mean(rating)) %>% 
  plot_ly(x=year, y=Avg_Rating)
```


# Exercise 2

This time group by year and Drama. In the <span class="func">summarize</span> create average rating again, but also a variable representing the average number of votes. In your plotly line, use the size and color arguments to represent whether the average number of votes and whether it was drama or not respectively. Use `mode = 'markers'`.

```{r}
movies %>% 
  group_by(year, Drama) %>% 
  summarise(Avg_Rating=mean(rating),
            Avg_Votes = mean(votes)) %>% 
  plot_ly(x=year, y=Avg_Rating, size=Avg_Votes, color=Drama, mode='markers')
```


# Exercise 3

Create a ggplot of your design and then use <span class="func">ggplotly</span> to make it interactive.
